---
title: "Coming from the Stata command"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Coming from the Stata command}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 4)
library(slopepower)
```

## Who this is for

You have used `slopepower` in Stata -- the command of Nash et al. (2021) -- and
want to know what changes in R. This vignette is the map. It answers three
questions: which of your numbers will move (almost none, and the exceptions are
named), which of your keystrokes will move (rather more), and why each of those
changes was made rather than transcribing the original option for option.

If you have never used the Stata command, read
`vignette("introduction", package = "slopepower")` instead; nothing here is
needed to use the package.

The rule the port followed throughout was: reproduce Nash et al.'s
*mathematics* line for line, and rewrite the *architecture* the way the authors
would have with more time -- fixing Stata-idiom artifacts and validation gaps,
never the statistical method. Every difference below falls on the architecture
side of that line, with two exceptions flagged as such in "Where the numbers
differ, on purpose".

## The numbers agree

Before the differences: given the same inputs, the two implementations produce
the same answers, and this has been checked rather than assumed. A Stata licence
was used to generate 584 grid rows and 37 model fits across all three models,
every comparator rule, both `var_tte` branches, the Dawson--Lagakos dropout
weighting and the `effectiveness` rescaling. Feeding Stata's own fitted variance
components into `slope_params_manual()`, all 542 comparable rows match exactly:
maximum absolute difference in *N* of 0, power agreeing to 3.4e-15.

Running the fits in each language rather than sharing them, 534 of those 542
still match *N* exactly, and every remaining difference traces to REML
convergence -- the optimiser, not the algebra. That is the one place the port's
numbers can differ from Stata's at all, and it is a consequence of the next
point.

### Convergence settings are fixed, not an option

Stata exposes `iterate(16000)` and otherwise leaves `mixed` at its defaults.
There is no equivalent argument here. The settings behind every fit live in one
exported function:

```{r control}
str(slope_lme_control()[c("maxIter", "msMaxIter", "opt", "tolerance")])
```

More iterations than `nlme`'s defaults, `opt = "optim"`, and tighter tolerances,
chosen because the stock settings converged less precisely on the two-block
random-effects structure fitted for `healthy`. It is exported so the settings
behind a published number can be inspected and reproduced from outside the
package -- **not** so they can be overridden from inside it. The model this
package fits is fixed (see `?slope_params`), and a tunable optimiser would let
two runs of the same call disagree with nothing on the returned object to say
why.

## The translation table

| Stata | R |
|---|---|
| `obs` + `nocontrols` | `slope_params(y ~ t \| id, data)` |
| `obs` + `casecon(v)` | `slope_params(y ~ t \| id, data, healthy = v)` |
| `rct` + `treat(v)` | `slope_params(y ~ t \| id, data, treated = v)` |
| `schedule(1 2 3)` | `trial_design(c(0, 1, 2, 3))` -- baseline explicit |
| `schedule(1 2 3 4) scale(0.5)` | `trial_design(seq(0, 2, by = 0.5))` -- no `scale` |
| `dropouts(.05 .05)` | `trial_design(..., dropout = c(0.05, 0.05))` |
| *(no equivalent)* | `dropout_type = "cumulative"`, `dropout_rate()` |
| `power(0.8)` | `slope_sample_size(params, design, power = 0.8)` |
| `n(450)` | `slope_power(params, design, n = 450)` |
| `effectiveness(0.33)` | `effectiveness = 0.33` |
| `usetrt` | `target = "observed"` |
| `nocontvar` | `common_variance = TRUE` |
| `iterate(16000)` | *(none -- see `slope_lme_control()`)* |
| *(no equivalent)* | `origin = "none"`, `slope_params_manual()`, the grid functions |

Everything in that table is available in one call through the compatibility
wrapper `slopepower()` as well; see "Keeping your old calls working" below.

## One command becomes several functions

Stata has a single command that decides at run time what you asked for, from
which of `power()` and `n()` you gave, and which of `obs`/`rct`/`nocontrols`
you declared alongside the matching variable. That flexibility costs it roughly
a dozen hand-written consistency checks: both of `power` and `n`, neither of
them, both of `obs` and `rct`, neither, `nocontrols` with `rct`, a declared
model whose variable is missing, a supplied variable the declared model cannot
use, and so on.

The R port makes each of those states unwritable instead of checked.

**Sample size and power are separate functions.** `slope_sample_size()` takes
`power`; `slope_power()` takes `n`. Neither takes the other:

```{r two-functions, error = TRUE}
pars1 <- slope_params(sdmt ~ visit | id, data = slpower1)
design <- trial_design(c(0, 1, 2))
slope_power(pars1, design, n = 200, power = 0.8)
```

That is R's own argument matching, not a check the package wrote. The ambiguity
Stata resolves with a run-time branch is a type error at the call site.

**The model is inferred from the data, not declared alongside it.** There is no
`obs`/`rct`/`nocontrols` argument. The scenario simply *is* which grouping
argument you supplied:

```r
slope_params(y ~ t | id, data)                    # Stata's obs nocontrols
slope_params(y ~ t | id, data, healthy = case)    # Stata's obs casecon()
slope_params(y ~ t | id, data, treated = treat)   # Stata's rct treat()
```

Of Stata's six model-consistency checks, exactly one survives, because it is the
only inconsistent state you can still express:

```{r one-check, error = TRUE}
slope_params(sdmt ~ visit | id, data = slpower3, healthy = treat, treated = treat)
```

**Fields are named for what they are.** Stata names its return scalars after the
model-specific role -- `slope_controls`, `slope_cases`, `slope_untreated`,
`slope_treated` -- so one concept, "reference slope versus index slope", has four
names across three models, and you cross-reference the on-screen legend to know
which you are holding. The R objects use one pair throughout: `slope` for the
untreated/case/index group and `slope_comparator` for the healthy controls or
treated arm (`NA_real_` where there is none), with a separate `comparator` field
(`"none"`, `"healthy"`, `"treated"`) carrying the model identity instead of the
names doing it.

## Trial designs are objects, and richer ones

### Baseline is explicit

Stata's `schedule()` lists follow-up visits only; visit 0 is inserted silently.
`trial_design()` requires you to write it:

```{r baseline, error = TRUE}
trial_design(c(1, 2))
```

`schedule(1 2)` becomes `trial_design(c(0, 1, 2))`. The implicit convention is a
common source of off-by-one design errors, and Stata's users have to remember it
rather than see it.

### Real-valued visit times, so no `scale()`

Stata restricts `schedule()` to ascending integers and builds the marginal
covariance matrix by looping over integer positions, which is why `scale()`
exists: six-monthly visits are expressed as an integer schedule on a rescaled
axis. This port builds Σ directly at whatever real times you supply, so
fractional spacing is written down directly and `scale()` has no equivalent:

```{r spacing}
slope_sample_size(pars1, trial_design(seq(0, 2, by = 0.5)),
                  effectiveness = 0.33)$n
```

620, which is exactly what Stata returns for `schedule(1 2 3 4) scale(0.5)` on
the same fit -- the equivalence is pinned against Stata's own output.

### Cumulative dropout, and dropout as a rate

Stata accepts only incremental dropout fractions (the proportion whose *last*
attended visit is visit *i*). Converting an attrition curve to that form is left
to you. `trial_design()` will do it:

```{r cumulative}
trial_design(c(0, 1, 2, 3), dropout = c(0, 0.10, 0.15),
             dropout_type = "cumulative")
```

The stored `dropout` field is always incremental whichever form you supplied,
and the cumulative vector is checked for being non-decreasing and bounded by 1.

`dropout_rate()` goes further and has no Stata counterpart at all. "5% a year
over three years" has to be hand-converted to `dropouts(0.15)`,
`dropouts(.05 .05 .05)` or `dropouts(.025 ...)` depending on which schedule you
are comparing, with the arithmetic redone for every row of a comparison table.
`dropout_rate(0.05)` is expanded per schedule automatically, so one object drives
every row:

```{r rate}
tbl <- slope_power_grid(
  pars1, n = 450, effectiveness = 0.33,
  visits  = list(annual = 0:3, six_month = seq(0, 3, 0.5)),
  dropout = list(`5pc` = dropout_rate(0.05))
)
tbl[, c("design", "dropout", "n", "power")]
```

This is new surface area, but it computes exactly what a Stata user computes by
hand for `dropouts()`; it is not a change to the method.

## Fit once, evaluate many designs

Stata re-runs the whole `mixed` model for every row of a comparison table. Table
1 of the paper compares nine designs, and so performs nine REML fits of the same
data.

Here, stage one (`slope_params()`, one `nlme::lme` fit) and stage two (the
closed-form algebra given those variance components) are separate functions, so
the grid above fitted the model once and evaluated every design against the same
estimates. Beyond speed, it means the designs are compared on identical
parameters rather than on nine separately converged fits.

The separation goes further than convenience. Stage two has no notion of "the
data currently in memory" -- the whole Stata command is a `program` operating on
the loaded dataset, with `preserve`/`restore` to undo its own mutations.
`slope_var()`, `slope_sigma()`, `slope_effect_size()`, `slope_sample_size()`,
`slope_power()` and the grid functions are pure functions of a `slope_params`
object and a `trial_design` object. They can be called with no dataset loaded at
all:

```{r manual}
pars_published <- slope_params_manual(
  slope = -1.6, sigma2_intercept = 100, sigma2_slope = 2,
  sigma_cov = 5, sigma2_residual = 10
)
slope_sample_size(pars_published, design, effectiveness = 0.33)$n
```

`slope_params_manual()` is how you plan from variance components reported in a
paper -- or from Stata's own fitted numbers, which is how the exact-agreement
check above was performed. Stata has no such entry point; the in-memory model
fit is mandatory there.

## Undefined results are errors, not missing values

When the slope difference is zero, the sample size is undefined: Stata divides
by zero, returns a missing value, and completes as though nothing happened. The
R port says so:

```{r zero-diff, error = TRUE}
flat <- slope_params_manual(slope = -1.6, slope_comparator = -1.6,
                            comparator = "treated",
                            sigma2_intercept = 100, sigma2_slope = 2,
                            sigma_cov = 5, sigma2_residual = 10)
slope_sample_size(flat, design, target = "observed")
```

The same applies to a singular information matrix, a non-positive-definite
implied covariance, a non-PD random-effects matrix on a hand-built
`slope_params`, and to a missing `n` -- where Stata does not fail at all but
silently switches to solving for sample size at 80% power:

```{r missing-n, error = TRUE}
slope_power(pars1, design, n = NULL)
```

Supplying `effectiveness` alongside `target = "observed"` is an error too, where
`slopepower.ado` quietly overwrites it to 1:

```{r target-eff, error = TRUE}
pars3 <- slope_params(sdmt ~ visit | id, data = slpower3, treated = treat)
slope_sample_size(pars3, trial_design(c(0, 2, 3)),
                  target = "observed", effectiveness = 0.5)
```

What the two *report* afterwards still agrees: under `target = "observed"` the
result's `effectiveness` is `NA_real_`, which is Stata's missing value faithfully
preserved rather than a divergence.

**These are additions; most guards are not.** Stata already errors on `alpha`
outside (0,1), `power` outside (0,1), `effectiveness` outside (0,1], `n < 2` and
non-integer `n`. The R port restates those checks. It does not add them, and if
you have been relying on them you lose nothing.

## Advisories you can catch, and objects that stay valid

### Warnings are conditions

Stata prints every advisory with `display as error`, indistinguishable at the
call site from a genuine failure except by reading the text -- and a warning
fires again on every row of a hand-rolled grid loop. Two advisories here are
classed R conditions, so you can handle them by class rather than by
string-matching:

```{r condition}
withCallingHandlers(
  design_d <- trial_design(c(0, 1, 2), dropout = c(0.05, 0.05)),
  slopepower_baseline_dropout = function(w) {
    cat("caught:", class(w)[1], "\n")
    invokeRestart("muffleWarning")
  }
)
```

The other is `slopepower_tte_direction`. Only those two carry a class, and
deliberately: they are the ones a legitimate grid sweep is expected to trip on
most cells, so the grid functions collect them by class and report each once per
grid rather than once per design. Everything else is an ordinary `stop()` or
`warning()`.

### The time shift is a message, is optional, and is recorded

Stata always rebases each subject's time to start at zero and reports it as a
`WARNING` line among the genuine errors. The port does the identical shift by
default, but reports it with `message()` and records that it happened on the
returned object, so you can detect it programmatically instead of grepping
console output:

```{r shift}
d2 <- slpower2
d2$time <- as.numeric(d2$vdate) / 365
pars2 <- slope_params(sdmt ~ time | id, data = d2, healthy = case)
pars2$time_shifted
```

`origin = "none"` declines the shift altogether, which Stata has no way to
express. The default, `origin = "subject"`, is Stata's behaviour.

### `common_variance` has three states, not two

Stata's `nocontvar` is a bare flag: absent fits the full unstructured
random-effects block for controls, present reduces it, and if the full structure
fails to converge you read `"Model failure. Try using the nocontvar option"` and
retype the command. `common_variance` takes `NULL`, `TRUE` or `FALSE`, of which
the flag expresses the last two. The default, `NULL`, has no Stata equivalent:
try the full structure, fall back to the reduced one with a `message()` if it
fails, and record which was used:

```{r common-var}
pars2$common_variance
```

### Invariants are re-checked at every use

Stata validates once, in the `syntax` block, coupled to that call's arguments;
there is no persisted object to revalidate. Here `slope_params` and
`trial_design` are real objects whose invariants -- positive-definite Σ and *G*,
dropout length and total, finite visits -- are re-checked at the boundary of
every stage-two call, not only where the constructors ran. That closes the gap
for objects the constructors never saw: a hand-assembled `slope_params` with a
non-PD random-effects matrix masked by a large residual variance would otherwise
return a plausible, wrong, large *N*, and a `trial_design` whose `$dropout` was
edited after construction would escape the baseline-dropout warning entirely.

### Variance components are extracted by name

This one is invisible unless it breaks. Stata pulls fixed and random effects out
of `e(b)` by hard-coded column index, and the index shifts between the three
models and the `nocontvar` variant --
`exp(b[1,13] + b[1,14])^2` is the cases' residual variance in model 1, but
`exp(b[1,11] + b[1,12])^2` in the `nocontvar` branch. That is correct (it was
verified under a real licence, base log-SD plus log-ratio offset), but correct
only because the numbering is memorised rather than derived.

The port extracts every component from the fitted object by name -- the
random-effects block from `nlme::getVarCov()`'s dimnames, the residual from the
`varStruct` coefficients looked up by level name. Adding or reordering a term
cannot silently mis-map a component. It also sidesteps a trap on the R side of
the same convention: `nlme::varIdent` puts the *reference* level's SD in
`sigma(fit)`, and since the reference here is `control`, a naive `sigma(fit)^2`
returns the controls' residual variance instead of the cases' -- a 3.3% error in
the fits used here.

### The bootstrap has nothing to misconfigure

The paper's bootstrap recipe (§2.6, §4.1.2) is not built into the Stata command:
you assemble `cluster()`, `idcluster()`, `strata()` and `jack()` yourself, with a
caveat that it assumes no observations were excluded. Four things to get wrong,
with no structural guard. `slope_bootstrap()` has no equivalent options at all:
subjects are always the resampling unit, replicate identifiers are always
regenerated (so a subject drawn twice counts as two people rather than
colliding), and stratification by group is automatic whenever
`comparator != "none"`. It also applies the paper's own §2.6 recommendation --
warning when the fitted slope is under 2.5 times its standard error -- which the
Stata side leaves you to check by eye.

## Keeping your old calls working

`slopepower()` is the compatibility wrapper: one bimodal command with Stata's
argument names, Stata's defaults (`power` defaults to 0.8), Stata's model
declaration and every one of Stata's consistency checks, because accepting Stata
calls verbatim is its whole job.

```{r wrapper}
res <- slopepower(slpower1, "sdmt", "id", "visit", schedule = c(1, 2),
                  obs = TRUE, nocontrols = TRUE, effectiveness = 0.33)
res$n
```

Note `schedule = c(1, 2)` with the implicit baseline, exactly as in Stata. It
accepts `scale` too, though it will suggest you drop it:

```{r wrapper-scale}
slopepower(slpower1, "sdmt", "id", "visit", schedule = 1:4, scale = 0.5,
           obs = TRUE, nocontrols = TRUE, effectiveness = 0.33)$n
```

The warn-and-reset behaviour is preserved as well -- an option that does not
apply to the declared model warns and is ignored, rather than erroring:

```{r wrapper-warn}
res2 <- slopepower(slpower3, "sdmt", "id", "visit", schedule = c(2, 3),
                   obs = TRUE, nocontrols = TRUE, treat = "treat",
                   effectiveness = 0.33)
```

Internally that family of warnings is one shared helper rather than three
copy-pasted blocks, so the "does this option apply to this model" logic and its
wording live in one place. (One of the Stata originals contains an unbalanced
macro quote mid-string. Stata prints the backtick literally and carries on --
checked, `_rc = 0`, identical result -- so the wrapper's warn-and-continue is a
faithful reading, not a repair. The only difference is the stray character.)

The wrapper is for porting existing calls. New code should use the two-stage
interface, which is why the migration table above points at it.

## Where the numbers differ, on purpose

Two places, both bounded and both documented.

### `var_tte` under dropout at saturated power

With dropout, no single *s*\*² applies across strata, so `var_tte` is back-solved
by inverting the sample-size formula. Stata uses that one expression in both
directions -- when `power` is your input and when `power` is the number it just
computed. The port keeps Stata's form when solving for *N*, rounding included,
and uses an algebraically identical but numerically stable form when solving for
power.

The two agree exactly, because `n_per_arm` cancels out of the ratio. Round-tripping
through the power anyway costs precision, and past a few thousand per arm costs
correctness: power saturates at exactly 1, and the inverse normal of 1 is not a
number in either language, so Stata's expression returns a missing `var_tte` for
a calculation that otherwise succeeded. (An earlier literal transcription here
reported 0, `qnorm(1)` being `Inf`.) The stable form still reports:

```{r saturated}
pw <- slope_power(pars1, trial_design(c(0, 1, 2), dropout = c(0, 0.1)),
                  n = 10000, effectiveness = 0.33)
c(power = pw$power, var_tte = pw$var_tte)
```

When solving for *N* there is no such cancellation -- `n_per_arm` has been
rounded up to a whole participant -- so that branch keeps Stata's literal form
and reports the same very slightly inflated number Stata does.

### The published 1,272

Section 4.1.3 of the paper notes that powering for a fraction *p* of a
previously observed effect scales *N* by *p*<sup>-2</sup>, and quotes 1,272 for
*p* = 0.5 by multiplying 318 by 4. That multiplies an already-`ceiling()`-rounded
*N*, which magnifies the rounding. Rounding last gives 1,266:

```{r published}
ss3 <- slope_sample_size(pars3, trial_design(c(0, 2, 3), dropout = c(0.2, 0.1)),
                         target = "observed")
p <- 0.5
c(paper   = ss3$n * 4,
  rounded_last = 2 * ceiling((qnorm(1 - ss3$alpha / 2) + qnorm(ss3$power))^2 /
                             (p * ss3$effect_size)^2))
```

The test suite pins *both* numbers and asserts the gap is always upward and
bounded by *p*<sup>-2</sup> per arm. The paper's figure is not silently
corrected, and not silently accepted as the right answer either. Recruiting 1,272
is not wrong, merely a rounding either way from the same calculation.

## Things that look like differences and are not

Recorded because they are the ones a careful reader re-derives otherwise. Each
was settled by *running* Stata, not by reading it harder.

* **The dropout-total guard.** Stata accumulates by repeated subtraction, and
  `1 - 0.3 - 0.3 - 0.4` is `-5.6e-17` in double precision, so
  `dropouts(0.3 0.3 0.4)` looks certain to trip the "total exceeds 1" guard. It
  does not: a Stata local round-trips through a decimal *string*, so `1 - 0.3`
  stores as `.7` and the residue lands on exactly 0. Stata accepts it and returns
  N = 1418, which is what this port returns. The port's `1e-8` tolerance
  *restates* that behaviour rather than repairing it -- it is still the right way
  to write the check in R, where `sum()` does carry the full double, but it fixes
  nothing, and Stata's guard still fires correctly on a total genuinely over 1.
* **`dropouts()` is validated before the fit in Stata too.** The length and total
  guards sit in the syntax block, well before the `mixed` call. The port briefly
  had this backwards and it was corrected -- restoring Stata's ordering, not
  improving on it.
* **The `sched_length` macro-name space is harmless.** Two lines of the `.ado`
  write a trailing space inside a macro name. Stata trims it; both counters are
  correct and the length guard fires in both directions.
* **`exp(b13 + b14)^2` really is the cases' residual variance**, as described
  above, not two free log-SDs.

## Two things deliberately not added

Not omissions by oversight -- checked against the paper and left out, because
adding them would go beyond what Nash et al. specify.

* **Unequal treatment allocation.** The two-person treatment-effect-variance
  construction (one notional participant per arm) depends on 1:1 allocation;
  generalising it is not implied by anything the paper computes.
* **Intermittent (non-monotone) missingness.** Both implementations assume
  dropout is monotone -- once a participant misses a visit they are gone. The
  Dawson--Lagakos pattern-mixture approach the paper uses covers monotone dropout
  only, and a participant who misses one visit and returns has no representation
  in either language.

## Where to go next

* `vignette("introduction", package = "slopepower")` -- the package on its own
  terms, with the three worked examples from section 4 of the paper.
* `?slope_params` -- the three models in full, including what they deliberately
  do not support.
* `vignette("harmful-previous-trial", package = "slopepower")` -- what to reuse
  from a trial in which the treatment made things worse, and why
  `target = "observed"` is the wrong tool there.
* `vignette("what-is-s-star", package = "slopepower")` -- what `s*` actually is
  in both implementations, why the `s*/sqrt(N)` scaling is exact, and why 1:1
  allocation and monotone dropout are limits of the construction rather than of
  the port.
* `DIVERGENCES.md` in the package sources -- the same ground as this vignette,
  written for someone auditing the port against the `.ado` file line by line,
  with the Stata source of each item quoted.

## Reference

Nash, S., Morgan, K. E., Frost, C., & Mulick, A. (2021). Power and
sample-size calculations for trials that compare slopes over time:
Introducing the `slopepower` command. *The Stata Journal*, 21(3), 575–601.
<https://doi.org/10.1177/1536867X211045512>
