---
title: "Reusing a trial in which the treatment did harm"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Reusing a trial in which the treatment did harm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 4)
library(slopepower)
```

## The situation

You ran a two-arm trial of a treatment meant to slow disease progression. It did
the opposite: the treated arm's outcome declined *faster* than the control arm's.
You now want to plan the next trial. What, if anything, can you take from the one
you have?

The short answer is: the **variance components** and the **control-arm slope**,
but *not* the treatment effect. This vignette explains why `target = "observed"`
is the wrong tool here, sets out the two ways to reuse the data that are
defensible, and then measures how good the resulting estimates actually are.

Throughout we use a simulated trial built from the generating model of Nash et
al. (2021, §4.1.3) — the one behind the packaged `slpower3` — except that the
treated arm is made worse than the control arm rather than better:

```{r generator}
truth <- list(
  intercept = 34,
  slope_control = -1.8,   # units/year
  slope_treated = -2.8,   # harm: treated decline faster
  s2a = 100, s2b = 2, sab = 5, s2e = 10
)
visits_observed <- c(0, 0.5, 2)   # visit times in the trial we already ran

sim_trial <- function(n_arm = 75, tr = truth, tvis = visits_observed) {
  n  <- 2 * n_arm
  G  <- matrix(c(tr$s2a, tr$sab, tr$sab, tr$s2b), 2, 2)
  ab <- matrix(rnorm(2 * n), n, 2) %*% chol(G)
  trt <- rep(0:1, each = n_arm)
  sl  <- ifelse(trt == 1, tr$slope_treated, tr$slope_control)
  d <- data.frame(
    id    = rep(seq_len(n), each = length(tvis)),
    treat = rep(trt,        each = length(tvis)),
    visit = rep(tvis,       times = n)
  )
  d$sdmt <- tr$intercept + sl[d$id] * d$visit +
    ab[d$id, 1] + ab[d$id, 2] * d$visit + rnorm(nrow(d), 0, sqrt(tr$s2e))
  d
}

set.seed(20260805)
trial <- sim_trial()
```

Stage one recovers what happened:

```{r fit}
pars <- slope_params(sdmt ~ visit | id, trial, treated = treat)
pars
```

The `observed difference in slopes` is control minus treated, so a **positive**
value means the treatment harmed. The proposed trial, throughout, is three years
with an interim visit at year two:

```{r design}
design <- trial_design(c(0, 2, 3))
```

## Why `target = "observed"` is the wrong tool

`target = "observed"` (Stata's `usetrt`) takes the observed treatment effect as
the target for the next trial. It is defined only for a treatment that helped.
Nothing stops you calling it here, and that is the problem: the sample-size
formula is

$$N \;\propto\; \frac{s_*^2}{\beta_2^2}$$

and the target effect \(\beta_2\) enters **squared**. The sign cancels. A trial
powered to detect a benefit of 1.0 units/year and a trial powered to detect a
*harm* of 1.0 units/year get the same answer.

To see it, run the same data twice — once as it came out, and once with the arm
labels swapped, which is the counterfactual trial where the same treatment helped
by the same amount:

```{r sign-cancels}
flipped <- transform(trial, treat = 1L - treat)
pars_flipped <- slope_params(sdmt ~ visit | id, flipped, treated = treat)

harm    <- suppressWarnings(slope_sample_size(pars, design, target = "observed"))
benefit <- slope_sample_size(pars_flipped, design, target = "observed")

data.frame(
  scenario    = c("treatment harmed", "treatment helped (labels swapped)"),
  slope_ctrl  = c(pars$slope, pars_flipped$slope),
  slope_treat = c(pars$slope_comparator, pars_flipped$slope_comparator),
  tte         = c(harm$tte, benefit$tte),
  n           = c(harm$n, benefit$n)
)
```

Identical `n`. Only the sign of the target treatment effect differs, and in the
harm row it is negative — meaning the design is powered to detect a treatment
that makes patients *worse* by that amount. That is a coherent calculation and
almost never the trial anyone wants to run.

`slopepower` warns rather than refusing, because the calculation is not wrong,
only misdirected:

```{r warn}
tryCatch(slope_sample_size(pars, design, target = "observed"),
         warning = function(w) conditionMessage(w))
```

The Stata original prints the negative target silently and moves on
(`slopepower.ado`, lines 449–457), so if you are checking an old script against
this package, expect the warning to be new.

Note also that `effectiveness` cannot express harm at all: it is a proportion of
a gap that treatment closes, constrained to \((0, 1]\) by both implementations.
There is no way to ask for a harmful target except through `target = "observed"`.

## What is actually reusable

Two things, and neither is the treatment effect:

* the **variance components** \(\sigma^2_a, \sigma^2_b, \sigma_{ab}, \sigma^2_\varepsilon\)
  (`sigma2_intercept`, `sigma2_slope`, `sigma_cov`, `sigma2_residual`), which
  govern \(s_*\) and are the parameters the paper describes as hardest to guess
  a priori (§2.2);
* the **control-arm slope**, which is the untreated rate of progression in your
  population.

Both are direction-agnostic. A trial that failed still measured them.

The target treatment effect for the next trial then has to come from clinical
judgment, expressed as an `effectiveness` fraction of that control-arm slope.

## Two defensible options

### A. Keep both arms, take the target from the control arm

Fit the previous-trial model — one intercept, arm-specific slopes, shared
variance components — but supply `effectiveness` instead of
`target = "observed"`. The observed effect is then discarded entirely, while all
150 participants contribute to the variance components:

```{r option-a}
opt_a <- slope_sample_size(pars, design, effectiveness = 0.25)
opt_a
```

### B. Control arm only

Drop the treated arm and treat what remains as a single untreated cohort:

```{r option-b}
control_only <- subset(trial, treat == 0)
pars_b <- slope_params(sdmt ~ visit | id, control_only)
opt_b  <- slope_sample_size(pars_b, design, effectiveness = 0.25)
opt_b
```

The paper endorses this route explicitly for a previous trial of an unrelated
treatment (p. 594): *"we might have decided to use only the untreated arm as our
dataset and use the options for a single group of untreated subjects."*

### Choosing between them

```{r compare-ab}
rbind(cbind(option = "A: both arms",   as.data.frame(opt_a)),
      cbind(option = "B: control only", as.data.frame(opt_b)))[
        , c("option", "n", "tte", "var_tte", "slope", "n_subjects")]
```

Do not read anything into which of the two gives the smaller `n`. On this draw
option B does, by a wide margin, and that is sampling noise: `var_tte` is
estimated from 75 participants instead of 150, so it is simply a noisier number,
and here it happened to come out low. Option A uses twice the data for the
variance components and is therefore the *more precise* of the two, even when it
returns the larger sample size. Choosing between them on the basis of the answer
you prefer is a good way to under-power the next trial.

Option A buys that precision with an assumption: the previous-trial model
constrains \(\sigma^2_a, \sigma^2_b, \sigma_{ab}, \sigma^2_\varepsilon\) to be
**equal in both arms**.

In this particular setting that assumption is the weak point. You are pooling the
variability of people the treatment harmed into an estimate you intend to apply to
a future *control* arm. A treatment that changes the mean trajectory can
plausibly change the spread around it too — more heterogeneous response, or
noisier measurement in a group doing worse. If you suspect that, option B trades
precision for a cleaner target and is the better choice. Comparing the two fits
is a cheap diagnostic: a large discrepancy in the variance components is evidence
against pooling.

Option A also forces a **common intercept** across arms, which is the
randomisation assumption of the analysis model (§2.1). Over repeated
randomisations that is fine; conditional on a realised chance baseline imbalance,
that imbalance leaks into the arm-specific slopes. Option B, having a free
intercept, is immune.

## Are these estimates unbiased?

Stage one is. The sample size that comes out of stage two is not — and the reason
is worth understanding, because it tells you when it matters.

The stage-one estimates are REML fits of a correctly specified linear mixed
model. The fixed effects are unbiased under the usual conditions, and REML exists
precisely to remove the downward bias that maximum likelihood has in variance
components. But

$$N \;\propto\; \frac{s_*^2}{(e\,\hat\beta_1)^2}$$

puts an estimated slope in the denominator, squared. \(1/x^2\) is convex, so
Jensen's inequality guarantees \(\mathbb{E}[1/\hat\beta_1^2] > 1/\beta_1^2\): the
sample size is biased **upward** even though every input is unbiased. A
second-order Taylor expansion quantifies it in terms of the signal-to-noise ratio
of the slope, \(r = |\beta_1| / \mathrm{SE}(\hat\beta_1)\):

$$\mathbb{E}\!\left[\frac{1}{\hat\beta_1^{2}}\right] \;\approx\;
  \frac{1}{\beta_1^{2}}\left(1 + \frac{3}{r^{2}}\right)$$

### Measuring it

Because we generated the data, the truth is available: put the true parameters
through stage two and compare.

```{r truth}
true_pars <- slope_params_manual(
  slope = truth$slope_control, sigma2_intercept = truth$s2a,
  sigma2_slope = truth$s2b, sigma_cov = truth$sab,
  sigma2_residual = truth$s2e
)
true_res <- slope_sample_size(true_pars, design, effectiveness = 0.25)
c(n = true_res$n, var_tte = true_res$var_tte)
```

Now repeat the whole pipeline — option A, no dropout — on fresh datasets:

```{r replicate}
R_sim <- 300   # the figures quoted in the text come from 2000

set.seed(1)
out <- t(vapply(seq_len(R_sim), function(i) {
  d <- sim_trial()
  p <- slope_params(sdmt ~ visit | id, d, treated = treat)
  z <- suppressWarnings(slope_sample_size(p, design, effectiveness = 0.25))
  c(slope = p$slope, s2a = p$sigma2_intercept, s2b = p$sigma2_slope,
    sab = p$sigma_cov, s2e = p$sigma2_residual, var_tte = z$var_tte, n = z$n)
}, numeric(7)))

truths <- c(slope = truth$slope_control, s2a = truth$s2a, s2b = truth$s2b,
            sab = truth$sab, s2e = truth$s2e,
            var_tte = true_res$var_tte, n = true_res$n)

summary_tab <- data.frame(
  truth    = truths,
  mean     = colMeans(out),
  median   = apply(out, 2, median),
  bias_pct = 100 * (colMeans(out) - truths) / abs(truths),
  mcse_pct = 100 * apply(out, 2, sd) / sqrt(nrow(out)) / abs(truths)
)
round(summary_tab, 3)
```

Every stage-one input sits on its true value to within Monte Carlo error, and so
does `var_tte`. The sample size does not: it is biased upward by
`r sprintf("%.1f%%", summary_tab["n", "bias_pct"])` here, with a Monte Carlo
standard error of `r sprintf("%.1f", summary_tab["n", "mcse_pct"])` percentage
points. Run at 2000 replicates instead of `r R_sim`, this same setup gives
**+8.6%** (MCSE 1.1%).

The delta-method formula predicts that from the slope's signal-to-noise ratio
alone:

```{r delta-check}
r_hat <- abs(truth$slope_control) / sd(out[, "slope"])
c(signal_to_noise = r_hat,
  predicted_bias_pct = 100 * 3 / r_hat^2,
  observed_bias_pct  = summary_tab["n", "bias_pct"])
```

### Where the bias comes from

Recomputing the per-arm sample size with only one of the two inputs estimated
isolates the source. (No dropout here, so `var_tte` is the plain \(s_*^2\); with
dropout the field holds a different quantity.)

```{r decompose}
k  <- (qnorm(0.975) + qnorm(0.8))^2
e  <- 0.25
n_true <- true_res$var_tte / (e * truth$slope_control)^2 * k

data.frame(
  estimated = c("nothing (truth)", "slope only", "variances only", "both"),
  per_arm_n = c(n_true,
                mean(true_res$var_tte / (e * out[, "slope"])^2 * k),
                mean(out[, "var_tte"] / (e * truth$slope_control)^2 * k),
                mean(out[, "var_tte"] / (e * out[, "slope"])^2 * k))
) |> transform(bias_pct = 100 * (per_arm_n - n_true) / n_true)
```

Essentially all of it is the slope in the denominator; the variance components
contribute almost nothing. The two contributions simply add, which is what you
would expect given that fixed effects and variance components are asymptotically
orthogonal in a Gaussian linear mixed model.

### The bias is in the safe direction, and the estimator is median-unbiased

```{r median-unbiased}
c(median_n = median(out[, "n"]), true_n = true_res$n,
  P_below_truth = mean(out[, "n"] < true_res$n))
```

The mean is pulled up by a long right tail; the median — which is what you
actually report — sits on the truth to within Monte Carlo error. (At `R_sim` =
`r R_sim` the median is itself noisy; the 2000-replicate run gives a median of
636 against a truth of 640, and `P_below_truth` = 0.505.) The bias that remains
is conservative: you over-recruit rather than under-recruit.

## Does the bias matter? Three things matter more

### 1. The spread dwarfs it

```{r spread}
q <- quantile(out[, "n"], c(.025, .1, .25, .5, .75, .9, .975))
round(q)
```

The 95% range spans a factor of about `r sprintf("%.1f", q[["97.5%"]] / q[["2.5%"]])`
against a truth of `r true_res$n`, and that is with a well-determined slope
(\(r \approx `r sprintf("%.1f", r_hat)`\)). A ten-per-cent bias in the mean is
noise against that.

This is why §2.6 and §4.1.2 of the paper press you to bootstrap the whole
pipeline rather than quote a point estimate:

```{r bootstrap}
set.seed(2)
slope_bootstrap(slope_sample_size(pars, design, effectiveness = 0.25),
                R = 120, type = "bca")
```

Use a much larger `R` in earnest — the paper uses 2000. Note that the bias
scales as \(3/r^2\), so the paper's rule of thumb in §2.6 that the slope should
exceed 2.5 times its standard error is, read this way, a bound on exactly this
bias: at \(r = 2.5\) it is about 48%, at \(r = 6\) about 8%, at \(r = 10\) about
3%.

### 2. The `effectiveness` you choose dominates everything

\(N \propto e^{-2}\), so:

```{r effectiveness-sensitivity}
es <- c(0.15, 0.20, 0.25, 0.33, 0.50)
data.frame(
  effectiveness = es,
  n = vapply(es, function(e) slope_sample_size(pars, design, effectiveness = e)$n, 1)
)
```

Moving from 0.25 to 0.33 changes `n` by a factor of \((0.33/0.25)^2 = 1.74\) —
larger than the estimation bias and larger than the interquartile range of the
sampling distribution. That number is a clinical judgment with no sampling
distribution at all, and the trial you already ran tells you nothing about it,
because the one quantity it estimated is the one you cannot use.

### 3. Specification bias, which no sample size fixes

The simulation above generates data from the model, so it can only speak to
estimation error. The biases that do not shrink as the previous study grows are:

* **Common variance across arms** (option A only) — discussed above, and the one
  most likely to bite when the treatment did harm.
* **Linearity.** The fitted slope is a projection of the true trajectory onto a
  straight line, and which line you get depends on the *previous* trial's visit
  schedule. Here we costed visits at years 2 and 3 from data ending at year 2.
  `slopepower` gives no warning when you extrapolate (§2.6). With a bounded
  outcome such as SDMT, floor effects make curvature likely.
* **Transportability.** §2.6 requires the previous population to resemble the
  future trial's. A trial that harmed people may also have had unusual entry
  criteria or an unusual case mix.
* **Missing-not-at-random dropout** in the previous trial flattens the observed
  slope and so inflates `n`. A harmful treatment is exactly the setting where
  differential withdrawal is likely.

None of these is visible in a bootstrap interval.

## Summary

1. Do **not** use `target = "observed"` on a trial that showed harm. It returns
   the same sample size it would have for a benefit of equal magnitude, and the
   target effect it reports is negative.
2. Reuse the variance components and the control-arm slope; take the target
   effect from clinical judgment via `effectiveness`.
3. Prefer **option A** (both arms, `effectiveness`) for precision; switch to
   **option B** (control arm only) if you doubt that the treatment left the
   variance structure alone. Compare the two fits — it costs one line.
4. The point estimate of `n` is median-unbiased and its mean is conservative by
   roughly \(3/r^2\). Do not correct for it.
5. Spend the effort instead on `slope_bootstrap()` and on a sensitivity analysis
   over `effectiveness` and over options A versus B. That is where the
   uncertainty actually lives.

## Reference

Nash, S., Morgan, K. E., Frost, C., & Mulick, A. (2021). Power and sample-size
calculations for trials that compare slopes over time: Introducing the
`slopepower` command. *The Stata Journal*, 21(3), 575–601.
<https://doi.org/10.1177/1536867X211045512>
