---
title: "An introduction to the slopepower package"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{An introduction to the slopepower package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 4)
library(slopepower)
```

## The idea in one sentence

You have a trial to plan whose outcome is not a single measurement but a
*rate of change* -- cognitive decline over years, say, rather than a score at
one follow-up visit. `slopepower` sizes that trial in two stages: first fit a
linear mixed model to some longitudinal data you already have, to learn how
fast people change and how much they vary in doing so; then combine that fit
with the visit schedule and dropout pattern of the trial you are planning to
get a sample size or a power. The two stages are two function calls,
`slope_params()` and then `slope_sample_size()` or `slope_power()`.

This vignette is a tour of the three situations that `slope_params()`
supports, following the structure of the worked examples in section 4 of
Nash et al. (2021), who wrote the Stata command this package ports. Each
situation is illustrated with one of the three datasets the package ships:
`slpower1`, `slpower2` and `slpower3`. All three are simulated, and all three
share an outcome: the Symbol Digit Modalities Test (SDMT), a measure of
cognitive processing speed in which lower scores are worse. The simulated
disease is a stand-in for a progressive condition such as Huntington's
disease, in which SDMT tends to fall over time.

## Three shapes of previous data

The previous data you have access to determines which arguments you pass to
`slope_params()`, and what a "100% effective" treatment is taken to mean:

* **Only untreated participants with the disease** (`slpower1`). There is no
  reference group at all, so the only sensible target for a fully effective
  treatment is to stop decline altogether -- a slope of zero.
* **Untreated participants plus a disease-free comparison group**
  (`slpower2`). Now a fully effective treatment is one that brings decline
  down to the rate seen in people who do not have the disease, so the target
  is the healthy group's slope rather than zero.
* **Both arms of a previous randomised trial** (`slpower3`). Here you already
  have an observed treatment effect, and can either aim to detect the same
  effect again or fall back to a target expressed as a fraction of decline
  removed, ignoring what the previous trial's treated arm did.

Each is a different call to `slope_params()`, distinguished by which of the
`healthy` and `treated` arguments (if either) you supply, and each returns an
object that carries a `comparator` telling the rest of the package which of
these three situations it came from.

## Example 1: only untreated participants (`slpower1`)

`slpower1` has 200 people with the simulated disease, each seen four times, a
year apart:

```{r slpower1-head}
subset(slpower1, id == 1)
```

### Fitting the previous data

```{r slpower1-fit}
pars1 <- slope_params(sdmt ~ visit | id, data = slpower1)
pars1
```

The estimated annual decline is -1.672 points. Alongside it are the four
variance components of the random-intercept, random-slope model: how much
people vary in where they start (`sigma2_intercept`), how much they vary in
how fast they decline (`sigma2_slope`), the covariance between the two, and
the residual measurement variance. `slope_sample_size()` and `slope_power()`
need all five numbers, not just the slope, which is why stage one returns an
object rather than a bare estimate.

### Sizing a trial

Suppose the planned trial runs two years with annual visits, and you want 80%
power (the default) to detect a treatment that removes a third of the
decline seen above:

```{r slpower1-design}
design1 <- trial_design(c(0, 1, 2))
design1
```

`trial_design()` always takes the baseline visit explicitly, at time 0 --
unlike Stata's `schedule()`, which lists follow-up visits only and assumes an
implicit baseline. Now the sample size:

```{r slpower1-ss}
ss1 <- slope_sample_size(pars1, design1, effectiveness = 0.33)
ss1
```

712 participants, 356 per arm -- the treatment difference `slopepower` is
powering the trial to detect, 0.552 points/year, is a third of the estimated
1.672-point decline.

### Longer follow-up can *shrink* the sample size

Extending the same trial to five years, with one extra visit at year five and
an assumed 10% loss to follow-up between years two and five, actually
*reduces* the sample size, even though some participants are now lost along
the way:

```{r slpower1-longer}
design1b <- trial_design(c(0, 1, 2, 5), dropout = c(0, 0, 0.1))
design1b
ss1b <- slope_sample_size(pars1, design1b, effectiveness = 0.33)
ss1b$n
```

A longer follow-up gives each participant's own trajectory more time to
diverge from the population average, which makes the individual slope easier
to estimate. That gain in precision outweighs the participants dropout
removes, at least for this much dropout and this much extra time -- it will
not always. `dropout` here has one entry fewer than `visits`; the print
method above shows how each entry maps onto "last attended visit" and
"first missed visit" so the pairing is never ambiguous.

### A different visit spacing

Stata's `schedule()` works on an integer grid and needs a `scale()` option to
express, say, six-monthly visits: you would give the schedule in half-year
units and set `scale(0.5)` to convert back to years. This port has no
`scale()` argument -- `trial_design()` takes real-valued visit times
directly, in whatever units the fitted slope is in:

```{r slpower1-spacing}
design1c <- trial_design(seq(0, 2, by = 0.5))
ss1c <- slope_sample_size(pars1, design1c, effectiveness = 0.33)
ss1c$n
```

620 participants: fewer than the annual-visit design, because the interim
visits again add precision, and *the same total* Stata would report for its
rescaled six-monthly version of this design. That equality is not a
coincidence -- it is a useful sanity check. The sample-size formula is a
ratio of a variance to a squared target difference, and rescaling time
multiplies both the numerator and the denominator by the same factor, so it
cancels. As long as the time units in `slope_params()` and `trial_design()`
agree with each other, it makes no difference what they are: years, months
or anything else all give the same `n`.

### How much should you trust that number?

A sample size computed from an estimated slope and four estimated variance
components is itself an estimate, with its own sampling uncertainty.
`slope_bootstrap()` resamples participants (not individual measurements),
refits stage one on each resample, and recomputes stage two, giving an
interval rather than a single number:

```{r slpower1-boot}
set.seed(42)
slope_bootstrap(ss1, R = 100, type = "bca", seed = 42)
```

`R = 100` is enough to illustrate the idea and keeps this vignette quick to
build; a bootstrap you intend to report from should use hundreds of
replicates, and this method's `type = "bca"` default over `"percentile"` is
deliberate, because the sampling distribution of a sample size is typically
right-skewed rather than symmetric.

## Example 2: untreated participants plus healthy controls (`slpower2`)

`slpower2` has 250 people with the disease and 250 disease-free controls,
each seen four times, with visit *dates* recorded rather than a visit
number:

```{r slpower2-head}
subset(slpower2, id %in% c(1, 251))
```

### Fitting the previous data

The time variable has to be built before fitting -- here, elapsed years
since a fixed reference date:

```{r slpower2-fit}
d2 <- slpower2
d2$time <- as.numeric(d2$vdate) / 365
pars2 <- slope_params(sdmt ~ time | id, data = d2, healthy = case)
pars2
```

Two things changed from example 1. First, `slope_params()` reported -- via a
`message()`, not a warning -- that time did not start at zero for everyone.
That is expected here, since `vdate` is a calendar
date, and by default (`origin = "subject"`) each participant's own first
visit is shifted to time zero, exactly as the Stata command does, so that the
random intercept is estimated at baseline rather than at an arbitrary shared
date. Second, the output now carries *two* slopes: -1.715 points/year for
the cases and 0.975 for the healthy controls, whose positive sign reflects a
practice effect in this simulated cohort rather than genuine improvement.
The variance components reported are always the cases', not the controls' --
see `?slope_params` for why that asymmetry is deliberate.

### Sizing a trial against the healthy-control slope

```{r slpower2-ss}
design2 <- trial_design(c(0, 1, 2))
ss2 <- slope_sample_size(pars2, design2, effectiveness = 0.33)
ss2
```

296 participants for the same 33% effectiveness and the same visit schedule
as the very first calculation in this vignette, and noticeably fewer than
the 712 there. The reference point moved: "33% effective" is now a third of
the *gap* between the cases' decline and the healthy controls' slope, 2.690
points/year, rather than a third of decline toward zero. Because the
controls are improving rather than merely stable, that gap is larger than
the cases' raw slope, so the same effectiveness fraction implies a bigger
target difference (0.888 against 0.552 earlier) and a smaller trial.

### Power for a fixed sample size, with dropout

If recruitment caps the trial at 200 participants rather than 296, the
question flips from "how many do I need" to "what power does 200 buy",
answered by `slope_power()` in place of `slope_sample_size()`. Here we also
assume 5% of those who start are lost each year:

```{r slpower2-power}
design2d <- trial_design(c(0, 1, 2), dropout = c(0.05, 0.05))
pw2 <- slope_power(pars2, design2d, n = 200, effectiveness = 0.33)
pw2
```

60% power -- noticeably short of the conventional 80% threshold, which is
exactly the kind of trade-off `slope_power()` exists to expose before a
trial is committed to a sample size it cannot reach.

The warning above is worth reading rather than dismissing: with the
`c(0, 1, 2)` schedule and a first dropout entry of 0.05, that entry describes
participants whose *only* recorded visit is baseline. They contribute no
information about a slope at all and are dropped from the calculation
entirely -- this port tells you so; the original Stata command does the same
thing silently. Nothing here is wrong, but it is worth knowing which
participants a nonzero `dropout[1]` refers to.

A bootstrap interval works the same way as in example 1 --
`slope_bootstrap(ss2, R = ..., type = "bca")` -- but is markedly slower here.
The two-group model is a single joint fit, but a larger one: 2,000
measurements on 500 participants, with a block-diagonal random-effects
structure and separate residual variances per group. On top of that, the
leave-one-subject-out jackknife that a BCa interval requires costs one such
fit per one of the 500 participants, so it is omitted from this
vignette in the interest of build time; budget a few minutes for it in
practice, or use `type = "percentile"` for a quicker, less accurate
alternative.

## Example 3: both arms of a completed trial (`slpower3`)

`slpower3` is 150 people already randomised, 75 to each arm, seen at
baseline, six months and two years:

```{r slpower3-head}
subset(slpower3, id %in% c(1, 76))
```

### Fitting the previous trial

```{r slpower3-fit}
pars3 <- slope_params(sdmt ~ visit | id, data = slpower3, treated = treat)
pars3
```

The control arm declined by 1.852 points/year and the treated arm by 1.104,
an observed difference of 0.747. Because both arms come from an actual
trial, there is a real treatment effect sitting in this fit -- unlike
examples 1 and 2, where "effectiveness" was always a hypothetical fraction
supplied by the user.

### Targeting the effect already observed

If the next trial should be powered to detect that same 0.747-point
difference again, use `target = "observed"` in place of `effectiveness` --
the equivalent of the Stata command's `usetrt` option. Here the trial runs to
three years with an interim visit at year two, and dropout of 20% by year
two and a further 10% by year three:

```{r slpower3-observed}
design3 <- trial_design(c(0, 2, 3), dropout = c(0.2, 0.1))
ss3 <- slope_sample_size(pars3, design3, target = "observed")
ss3
```

318 participants. Note that `effectiveness` must not be supplied alongside
`target = "observed"`: the observed difference fixes it at 1 by definition.

### Targeting a fraction of that effect

A trial this size is often a pilot, and a common worry about a pilot's
treatment effect is that it is optimistic. If the next, larger trial should
only be powered to detect half of what was observed, there is no need to
refit anything: the sample-size formula has the target difference squared in
its denominator, so a target that is some proportion `p` of the observed
effect needs the sample size multiplied by `p^-2`. Section 4.1.3 of the paper
does exactly this for `p = 0.5`, multiplying the 318 above by 4:

```{r slpower3-fraction}
ss3$n * 4
```

1,272 participants, which is the figure the paper quotes. It is very slightly
conservative, and worth knowing why. The scaling law is exact on the
*unrounded* per-arm requirement, but `n` has already been rounded up to a
whole participant per arm, and multiplying magnifies that rounding. Rounding
at the end instead gives 1,266:

```{r slpower3-fraction-exact}
p <- 0.5
2 * ceiling((qnorm(1 - ss3$alpha / 2) + qnorm(ss3$power))^2 /
            (p * ss3$effect_size)^2)
```

That is also what you would get by passing `target = "observed"` into a fresh
calculation whose reference slope had already been halved, which
`slope_params_manual()` can build -- so 1,266 is the figure internally
consistent with the rest of the package, and 1,272 the published one. The gap
is bounded by `p^-2` participants per arm, so it is six people here and never
more than a handful; recruiting the paper's 1,272 is not wrong, merely a
rounding either way from the same calculation.

### If the previous trial is not a precedent for this one

`target = "observed"` only makes sense when the *next* trial is aimed at
the *same* treatment as the one already tested. If `slpower3` had instead
been a trial of some unrelated treatment, the honest use of it is to treat
its control arm as a single group of untreated subjects and fall back to
example 1's approach, fitted to `subset(slpower3, treat == 0)`. The package
vignette `vignette("harmful-previous-trial", package = "slopepower")` works
through this and related questions -- what can and cannot be salvaged from a
previous trial, and in particular why `target = "observed"` is the wrong
tool when that trial's treatment made things worse rather than better -- in
more depth than belongs here.

## Comparing many trial designs at once

Section 4.2 of Nash et al. (2021) asks a broader question than any single
call above: given a fixed recruitment target, how does power change across
several candidate visit schedules and several assumed dropout rates? Their
table 1 fixes a three-year trial recruiting 450 participants and compares
three schedules -- baseline and final visit only, annual visits, and
six-monthly visits -- against three dropout scenarios -- none, 5% a year and
10% a year -- using the `slpower1` fit from example 1. `slope_power_grid()`
does the whole comparison in one call:

```{r grid-power}
tbl <- slope_power_grid(
  pars1, n = 450, effectiveness = 0.33,
  visits  = list(final_only = c(0, 3), annual = 0:3, six_month = seq(0, 3, 0.5)),
  dropout = list(none = NULL, `5pc` = dropout_rate(0.05), `10pc` = dropout_rate(0.10))
)
tbl[, c("design", "dropout", "n", "power")]
```

Two more-visits-is-better patterns are visible here. Adding follow-up visits
raises power even with no dropout at all -- from about 80% with a single
follow-up visit to nearly 87% with six-monthly visits -- for the precision
reasons already seen in example 1. And the benefit of extra visits grows as
dropout gets worse, because a participant who withdraws still contributes
whatever interim measurements they did attend: with 10% annual dropout, the
six-monthly design nearly recovers the no-dropout power of the sparsest
schedule.

`dropout_rate()` is what makes the three schedules comparable on "the same"
dropout assumption despite having different numbers of visits: "10% per
year" is a different vector of per-visit proportions for a two-visit
schedule than for a five-visit one, and `dropout_rate()` expands a single
stated rate correctly for each.

`slope_sample_size_grid()` is the converse table: instead of fixing `n` and
letting power vary, it fixes the target power and reports the `n` each
combination needs.

```{r grid-n}
tbl_n <- slope_sample_size_grid(
  pars1, power = 0.8, effectiveness = 0.33,
  visits  = list(final_only = c(0, 3), annual = 0:3, six_month = seq(0, 3, 0.5)),
  dropout = list(none = NULL, `5pc` = dropout_rate(0.05), `10pc` = dropout_rate(0.10))
)
tbl_n[, c("design", "dropout", "n", "power")]
```

Read together, the two tables are the same design space from opposite ends:
fix a recruitment target and ask what power it buys, or fix a power target
and ask what recruitment it costs.

Both tables so far vary the trial and hold the assumptions fixed. The
assumptions can be varied in exactly the same way: `effectiveness`, `alpha`
and whichever of `n` and `power` the grid is not solving for each take a
list of values, and every one of them is priced against every design. The
uncomfortable question about the table above is how much of it depends on
assuming a treatment that slows decline by 33%, and a grid answers that
without leaving the table:

```{r grid-sensitivity}
tbl_s <- slope_sample_size_grid(
  pars1, visits = list(annual = 0:3), dropout = dropout_rate(0.05),
  power         = c(`80pc` = 0.8, `90pc` = 0.9),
  effectiveness = list(`25pc` = 0.25, `33pc` = 0.33, `40pc` = 0.4)
)
tbl_s[, c("design", "dropout", "effectiveness", "power", "n")]
```

The sample size scales with the inverse square of the effect being targeted,
so an assumed effectiveness the trial does not achieve is by far the most
expensive of the two axes to be wrong about -- much more so than the choice
between 80% and 90% power. Each axis reports its value in the column of the
same name; the names given to the levels are what identify a cell in any
error or warning message.

## Where to go next

* `?slope_params` sets out the three models fitted in stage one in full,
  including what they deliberately do not support -- covariate adjustment,
  further clustering, non-linear trajectories -- and why.
* `slope_params_manual()` builds a stage-one object by hand, for planning
  from numbers reported in a paper when no participant-level dataset is
  available to fit.
* `vignette("from-stata", package = "slopepower")` is a migration guide for
  existing Stata `slopepower` users, mapping arguments and results between
  the two implementations.
* `vignette("harmful-previous-trial", package = "slopepower")` is a deeper
  treatment of one case touched on above: what to reuse from a previous
  trial, and what not to, when its treatment effect cannot simply be
  targeted again.
* `vignette("what-is-s-star", package = "slopepower")` derives the single
  number all of the above rests on: what the "two-person trial" standard
  error is, why the `s*/sqrt(N)` scaling is exact, and the floor on sample
  size that no visit schedule can beat.

## Reference

Nash, S., Morgan, K. E., Frost, C., & Mulick, A. (2021). Power and
sample-size calculations for trials that compare slopes over time:
Introducing the `slopepower` command. *The Stata Journal*, 21(3), 575–601.
<https://doi.org/10.1177/1536867X211045512>
