---
title: "Serially correlated and unstructured residuals"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Serially correlated and unstructured residuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 4)
library(slopepower)
```

## The question

The model of Nash et al. (2021) gives each participant a random intercept and
a random slope, and treats everything else -- every visit's departure from the
participant's own straight line -- as independent noise with one variance,
$\sigma^2_\epsilon$. That is often a good approximation. It is a poor one when
departures persist: a participant who has a bad spell is still having it at the
next visit, so two measurements taken close together repeat each other rather
than each telling you something new.

The sample size cares, because the thing a planned schedule buys is
information about each participant's slope. Under independent residuals every
extra visit adds a fresh, independent measurement, and packing visits closer
together keeps paying. Under serial correlation it pays less. Planning a dense
schedule from the independent model can therefore promise a smaller trial than
the data will support.

`slope_params()` takes the residual structure the way `nlme` does, so if you
have written

```r
nlme::lme(y ~ time, random = ~ time | id, correlation = nlme::corAR1(form = ~ time | id))
```

you already know how to ask for it:

```r
slope_params(y ~ time | id, data, correlation = nlme::corAR1())
```

## The residual covariance

Writing $\epsilon_i$ for participant $i$'s vector of residuals, the default is
$\epsilon_i \sim N(0, \sigma^2_\epsilon I)$. With a structure it is
$N(0, R_i)$, with

$$
(R_i)_{jk} = \sigma^2_\epsilon \, \delta_j \delta_k \, \rho(t_{ij}, t_{ik}),
$$

in `nlme`'s notation: $\rho$ the correlation between two of the participant's
visits, and $\delta$ the visit-specific SD ratios of `varIdent()` (1 without
it). The structures supported, with $d = |s - t|$:

| Structure | `correlation =` | $\rho(s, t)$ | Stata `mixed` |
|---|---|---|---|
| independent (default) | `NULL` | 0 | `residuals(independent)` |
| AR(1) | `nlme::corAR1()` | $\phi^d$, integer times only | `residuals(ar 1, t())` |
| continuous-time AR(1) | `nlme::corCAR1()` | $\phi^d$ | `residuals(exponential, t())` |
| exponential | `nlme::corExp()` | $(1 - n)\, e^{-d / r}$ | |
| Gaussian | `nlme::corGaus()` | $(1 - n)\, e^{-(d / r)^2}$ | |
| unstructured | `nlme::corSymm()` | one per pair of visit times | |

Here $r$ is the range and $n$ the nugget, zero unless `nugget = TRUE`. Adding
`weights = nlme::varIdent(form = ~ 1 | time)`, with the time variable of the
formula as the stratum, gives each visit time its own variance; with
`corSymm()` that is the fully unstructured residual covariance, Stata's
`residuals(unstructured, t())`.

The `form` is the one thing set for you. A residual correlation here is always
within a participant and a function of the formula's time variable, because
that is what stage two needs in order to evaluate it at a schedule you are
planning. `nlme`'s own default for `corAR1()`, the order of observation,
miscounts the lag across a missed visit, so it is not used.

## An example with serial correlation

The data below are simulated: 300 participants seen every six months for three
years, with a random intercept and slope, a serially correlated residual that
halves in about nine months, and some pure measurement error on top -- the
classic random effects, serial correlation and measurement error model of
Diggle and colleagues.

```{r simulate}
set.seed(2026)
sim <- function(n = 300, visits = seq(0, 3, by = 0.5)) {
  G  <- matrix(c(100, 3, 3, 1.5), 2, 2)               # random intercept and slope
  ab <- matrix(rnorm(2 * n), n, 2) %*% chol(G)
  R  <- 12 * 0.4^abs(outer(visits, visits, "-")) +   # serial part, correlation 0.4 at 1 year
        diag(3, length(visits))                      # measurement error
  e  <- matrix(rnorm(n * length(visits)), n) %*% chol(R)
  y  <- 40 + ab[, 1] + outer(ab[, 2] - 1.5, visits) + e
  data.frame(id = rep(seq_len(n), each = length(visits)),
             years = rep(visits, n),
             score = as.vector(t(y)))
}
d <- sim()
```

Fit it twice: as the paper does, and with an exponential correlation and a
nugget, which is that model.

```{r fit}
independent <- slope_params(score ~ years | id, d)
serial <- slope_params(score ~ years | id, d,
                       correlation = nlme::corExp(nugget = TRUE))
serial
```

The fitted range of about 1.7 years is a correlation of $0.84 e^{-1/1.7}
\approx 0.47$ between visits a year apart, against 0.4 in the simulation. The
two fits are ordinary `nlme` objects, so they can be compared the usual way:

```{r aic}
AIC(independent$fit, serial$fit)
```

Now price five schedules over the same three years, from a baseline and a
final visit to monthly visits. The `truth` row uses the simulation's own
parameters, through `slope_params_manual()`, which takes the same constructors
with the values stated:

```{r schedules}
truth <- slope_params_manual(
  slope = -1.5, sigma2_intercept = 100, sigma2_slope = 1.5, cov_intercept_slope = 3,
  sigma2_residual = 15,
  correlation = nlme::corExp(c(-1 / log(0.4), 3 / 15), nugget = TRUE))

schedules <- list("0, 3" = c(0, 3), "yearly" = 0:3, "6-monthly" = seq(0, 3, 0.5),
                  "quarterly" = seq(0, 3, 0.25), "monthly" = seq(0, 3, 1 / 12))
per_arm <- sapply(list(independent = independent, serial = serial, truth = truth),
                  function(p) sapply(schedules, function(v)
                    slope_sample_size(p, v, effectiveness = 0.3)$n_per_arm))
per_arm
```

Read down the columns rather than across them -- each fit is from one sample
of 300, so the absolute numbers carry sampling error. Going from two visits to
monthly ones, the independent model promises the trial gets about
`r round(100 * (1 - per_arm["monthly", "independent"] / per_arm["0, 3", "independent"]))`%
smaller. The model with serial correlation promises
`r round(100 * (1 - per_arm["monthly", "serial"] / per_arm["0, 3", "serial"]))`%,
and the truth is
`r round(100 * (1 - per_arm["monthly", "truth"] / per_arm["0, 3", "truth"]))`%.
Thirty-seven visits a participant are being sold on the strength of independence
the data do not have.

The floor of the "What s\* is" vignette still applies under a serial
correlation, because the correlation dies away with the time between visits:
a long enough schedule still averages the residual out. It is just approached
more slowly, since dense visits repeat each other.

## An unstructured residual covariance

`corSymm()` estimates a separate correlation for every pair of visit times, and
with `varIdent()` a separate variance for every time, so it imposes no shape at
all. On `slpower1`, seen yearly for three years:

```{r unstructured}
un <- slope_params(sdmt ~ visit | id, data = slpower1,
                   correlation = nlme::corSymm(),
                   weights = nlme::varIdent(form = ~ 1 | visit))
un
```

The price of imposing no shape is that the structure exists only at the four
times it was estimated at. A schedule among them is fine; one that is not is
refused, because nothing was estimated about a visit at six months:

```{r unstructured-schedules, error = TRUE}
slope_sample_size(un, c(0, 1, 3), effectiveness = 0.33)$n
slope_sample_size(un, c(0, 0.5, 1), effectiveness = 0.33)
```

For the same reason it needs the participants to share a schedule: with visit
times recorded as dates, as in `slpower2`, every participant has their own
times, and there is nothing to estimate a correlation *between visit 2 and
visit 3* from. Use a structure that is a function of time there, such as
`corCAR1()` or `corExp()`.

There is a second price, which the printed object notes. An unstructured
correlation can absorb any covariance the random effects would otherwise
explain, so the random-intercept variance, the random-slope variance and the
residual variance are no longer separately identified -- the fit is one point
on a ridge of equally good decompositions. What *is* identified is the total
covariance they imply at the visit times, and that is all a sample size
depends on. For the same reason `slope_var_floor()`, which is built from the
random-effects variances alone, is refused.

## What stage two assumes

The sample size for a structured residual is the sample size for a trial
*analysed* with that structure: a mixed model whose residual covariance is
specified the same way. Every stage-two print says so in a note, as it does
for covariate adjustment. Planning with one structure and analysing with
another is a different calculation, and not one this package makes.

Two structures are refused outright. `corCompSymm()`, an exchangeable
correlation, adds the same covariance to every pair of a participant's visits,
which is exactly what the random intercept already does; the two cannot be
told apart. And `corAR1()` is defined by `nlme` on integer times only, so it
is refused on data with fractional times and for a planned schedule with them;
`corCAR1()` is the same correlation, $\phi^{|s - t|}$, at any times.
