---
title: "Adjusting for stratification factors"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Adjusting for stratification factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 4)
library(slopepower)
```

## The situation

You are planning the sample size for a trial from a randomised trial you
already ran. That trial stratified its randomisation on age group and sex, and
the new trial will do the same. Its analysis will then adjust for both, as
analyses of stratified trials should. This vignette shows how to make
`slope_params()` estimate the variance components that the adjusted analysis
will actually face, and why stratification makes that the right thing to do.

In short: pass the stratification factors as `covariates`, use the same
covariates the new trial's analysis model will use, and keep their
interactions with time.

## A previous trial, stratified on age group and sex

The data below are simulated. There are 240 participants in four strata (age
under or over 60, by sex), randomised in permuted blocks of four within each
stratum, and seen at years 0, 1 and 2. Being older and being male both lower
the baseline SDMT score and speed up its decline. That second effect is the
one that matters for a trial whose outcome is a slope.

```{r simulate}
sim_trial <- function(n = 240, visits = c(0, 1, 2)) {
  age_group <- factor(sample(c("<60", "60+"), n, TRUE), levels = c("<60", "60+"))
  sex       <- factor(sample(c("F", "M"), n, TRUE))
  stratum   <- interaction(age_group, sex)

  # Stratified randomisation: permuted blocks of 4 within each stratum
  treat <- integer(n)
  for (s in levels(stratum)) {
    idx <- which(stratum == s)
    blocks <- replicate(ceiling(length(idx) / 4), sample(rep(0:1, 2)))
    treat[idx] <- as.vector(blocks)[seq_along(idx)]
  }

  older <- age_group == "60+"
  male  <- sex == "M"
  G  <- matrix(c(80, 4, 4, 1.5), 2, 2)     # random intercept and slope
  ab <- matrix(rnorm(2 * n), n, 2) %*% chol(G)
  a  <- 45 - 6 * older - 3 * male + ab[, 1]
  b  <- -1 - 1.6 * older - 0.6 * male + 0.4 * treat + ab[, 2]

  d <- data.frame(id = rep(seq_len(n), each = length(visits)),
                  visit = rep(visits, n))
  d$age_group <- age_group[d$id]
  d$sex       <- sex[d$id]
  d$treat     <- treat[d$id]
  d$sdmt      <- a[d$id] + b[d$id] * d$visit + rnorm(nrow(d), 0, 3)
  d
}

set.seed(2026)
trial <- sim_trial()
```

Because of the stratification, the arms are balanced to within one participant
in every stratum:

```{r balance}
first <- !duplicated(trial$id)
with(trial[first, ], table(stratum = interaction(age_group, sex), treat))
```

## Stage one: unadjusted and adjusted

The previous trial had a treated arm, so we use the `treated` model, which
gives one intercept and a separate slope for each arm. First we fit it without
covariates, and then adjusted for the two stratification factors:

```{r fit}
unadj <- slope_params(sdmt ~ visit | id, trial, treated = treat)
adj   <- slope_params(sdmt ~ visit | id, trial, treated = treat,
                      covariates = ~ age_group + sex)
adj
```

`covariates` takes a one-sided formula of *baseline* variables. A variable
that changes within a participant is refused. Factors are dummy-coded, and
every column is centred over participants, so `slope` remains the slope of a
participant with this trial's average age and sex mix. By default
(`covariate_time = TRUE`), each covariate also enters interacted with time.
That interaction is what lets age and sex explain differences in *slope*, and
not just in baseline level.

The fitted `lme` object shows the full model, with the covariate columns
under internal names (`sp_cov_1` is the dummy for age group `60+` and
`sp_cov_2` is the dummy for `M`):

```{r fixef}
nlme::fixef(adj$fit)
```

The two `sp_time:sp_cov_*` terms estimate how much faster older and male
participants decline. The simulation used -1.6 and -0.6 points per year.

To see which part of the adjustment does the work, we also fit a version that
adjusts the baseline level only:

```{r compare}
base_only <- slope_params(sdmt ~ visit | id, trial, treated = treat,
                          covariates = ~ age_group + sex, covariate_time = FALSE)

pick <- function(p) unlist(p[c("slope", "slope_comparator", "sigma2_intercept",
                               "sigma2_slope", "sigma_cov", "sigma2_residual")])
round(rbind(unadjusted    = pick(unadj),
            baseline_only = pick(base_only),
            adjusted      = pick(adj)), 3)
```

The table shows three things:

* **The slopes barely move.** The arms were balanced on age group and sex, so
  the adjustment has almost nothing to correct in the treatment comparison. In
  a stratified trial, the adjustment changes how *precise* the estimates are,
  not what they estimate.
* **`sigma2_slope` falls only with the time interactions.** The variation in
  slopes that age group and sex explain leaves the random-slope variance and
  moves into fixed effects. Adjusting the baseline level alone shrinks
  `sigma2_intercept` but leaves `sigma2_slope` untouched.
* **`sigma2_residual` does not change.** Baseline covariates are constant
  within a participant, so they can explain only between-participant
  variation.

## Stage two: the sample size

The new trial runs for three years with annual visits and aims to slow decline
by 30%:

```{r design}
design <- trial_design(c(0, 1, 2, 3))

sizes <- lapply(list(unadjusted = unadj, baseline_only = base_only, adjusted = adj),
                slope_sample_size, design = design, effectiveness = 0.3)
data.frame(n_per_arm = vapply(sizes, `[[`, 1, "n_per_arm"),
           var_tte   = vapply(sizes, `[[`, 1, "var_tte"))
```

The fully adjusted fit gives the smallest sample size, because the variance of
the estimated slope is driven by `sigma2_slope` (see
`vignette("what-is-s-star")`). Adjusting the baseline level alone gains
nothing, and here it costs a little. The analysis model gives both arms one
intercept, so the baseline visit carries some information about each
participant's slope through the intercept–slope correlation. Explaining part
of the intercept weakens that correlation (from 0.41 to 0.26 here), so less
information is borrowed.

## Which fit to plan with

Plan with the stage-one model that matches the new trial's analysis model.

* **The covariates must match.** Stage two returns the variance of the
  treatment effect that the analysis will estimate. Variance components
  adjusted for age and sex describe an analysis adjusted for age and sex. If
  the new trial is analysed without the adjustment, these components overstate
  its power, and the unadjusted fit is the right one.
* **The time interactions must match too.** The outcome is a slope, so
  "adjusting for age and sex" helps only if the analysis model includes
  `age_group:time` and `sex:time` terms. Check the statistical analysis plan.
  If it adjusts the baseline level only, use `covariate_time = FALSE`, and
  expect almost no saving.
* **The coding must match.** `~ age_group + sex` adjusts for the two factors.
  If the analysis will adjust for the four strata as such, use
  `~ age_group * sex` instead. In either case, use the same age bands as the
  randomisation.

Stratification is what makes the adjusted plan safe to rely on. The sample
size formula has no term for chance imbalance between the arms. In an
unstratified trial, the adjusted analysis loses a little precision whenever
the arms happen to differ in age or sex. Stratifying on those same variables
removes that imbalance by design, so the adjusted variance is the variance
the trial will actually have. The converse also holds: a stratified trial
analysed *without* adjusting for its stratification factors has a
conservative test, and planning with the unadjusted fit then sizes the trial
for an analysis it should not be running.

## Caveats

* **Case mix.** Age group and sex change the slope here, so the centred
  `slope` depends on the previous trial's mix of strata. The target effect
  under `effectiveness` is a proportion of that slope. If the new trial will
  recruit a noticeably older or more male population, its average control-arm
  slope, and so its target effect, will differ.
* **Uncertainty.** `slope_bootstrap()` refits the adjusted model in every
  replicate, so a bootstrap interval for the adjusted sample size needs no
  extra arguments:
  `slope_bootstrap(slope_sample_size(adj, design, effectiveness = 0.3), R = 2000)`.
* **Small strata.** A factor level that a bootstrap resample leaves out is
  dropped from that refit rather than making it fail. With very small strata,
  the adjustment will vary from one replicate to the next.

## Where to go next

* `?slope_params`, section "Covariate adjustment", gives the model in full.
* `vignette("introduction", package = "slopepower")` tours the three stage-one
  scenarios.
* `vignette("what-is-s-star", package = "slopepower")` explains why
  `sigma2_slope`, rather than `sigma2_intercept`, drives the sample size.
