---
title: "What s* is, and where it comes from"
author: "slopepower"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{What s* is, and where it comes from}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "")
options(width = 88, digits = 6)
library(slopepower)
```

## The two sentences this vignette is about

Everything `slope_sample_size()` and `slope_power()` do rests on a single
number, \(s_*\). Section 2.2 of Nash et al. (2021) introduces it like this:

> Because the matrices in (5) can get very large, we will use a simplifying
> trick — we shall first calculate the treatment-effect standard error for a
> two-person trial \(s_*\). Because the standard error for the treatment effect
> from a trial with \(N\) independent subjects in each arm is \(s_*/\sqrt{N}\),
> it follows from standard theory that the sample size required [...]

That is the whole treatment. Two things are asserted and neither is shown:

1. that "the treatment-effect standard error for a two-person trial" is a
   well-defined quantity at all — a trial with one participant per arm cannot
   estimate any of \(\sigma_a^2, \sigma_b^2, \sigma_{ab}, \sigma_\epsilon^2\),
   so it is not obvious what standard error is meant; and
2. that the scaling is \(s_*/\sqrt{N}\), which is attributed to "standard
   theory" without saying which theory, whether the equality is exact or
   asymptotic, or what it needs in order to hold.

Section 2.4 then says only that the command "builds \(\Sigma\) for a two-person
trial using (3) [...] It then calculates the standard error of the treatment
effect for this two-person trial, \(s_*\), by using (5)."

This vignette fills that in. It works entirely in the two-person case, because
that turns out to be the whole story: everything else is one exact division.
Along the way \(s_*\) acquires a closed form the paper never states, which makes
several facts about trial design fall out immediately — including a floor on
sample size that no visit schedule, however long or however dense, can get
below.

Throughout, one participant's marginal covariance matrix at visit times
\(t = (t_0, t_1, \dots, t_m)^{\mathsf T}\) with \(t_0 = 0\) is

$$\Sigma_{jk} \;=\; \sigma_a^2 \;+\; t_j t_k \sigma_b^2 \;+\; (t_j + t_k)\sigma_{ab} \;+\; [\,j = k\,]\,\sigma_\epsilon^2,$$

which is `slope_sigma()`, and \(u = (1,1,\dots,1)^{\mathsf T}\).

```{r params}
pars <- slope_params_manual(
  slope = -1.8,
  sigma2_intercept = 100, sigma2_slope = 2,
  sigma_cov = 5, sigma2_residual = 10
)
visits <- c(0, 1, 2)
Sigma <- slope_sigma(pars, visits)
Sigma
slope_var(pars, visits)   # s*^2
```

## 1. What s* is not

Take the phrase "the treatment-effect standard error for a two-person trial"
literally: run a trial on two people, one per arm, fit the analysis model, read
off the standard error. You can do this. The result is not \(s_*\).

```{r literal, message = FALSE}
library(nlme)
set.seed(1)
two <- data.frame(id = factor(rep(1:2, each = 3)),
                  t  = rep(visits, 2),
                  g  = rep(0:1, each = 3))
two$y <- 34 - 1.8 * two$t + 0.6 * two$g * two$t + rnorm(6, 0, 3)

fit2 <- lme(y ~ t + g:t, random = ~ t | id, data = two, method = "REML")
summary(fit2)$tTable["t:g", c("Std.Error", "DF")]
```

Compare `sqrt(slope_var(pars, visits))` = `r round(sqrt(slope_var(pars, visits)), 4)`.
The two numbers are in the same neighbourhood, but that is luck. Look at what
the fit thinks the variance components are:

```{r literal-vc}
c(sigma2_intercept = as.numeric(VarCorr(fit2)[1, 1]),
  sigma2_slope     = as.numeric(VarCorr(fit2)[2, 1]),
  sigma2_residual  = as.numeric(VarCorr(fit2)[3, 1]))
```

against true values of 100, 2 and 10. With six observations and three fixed
effects there are two residual degrees of freedom and nothing to estimate a
covariance structure from; the reported standard error is an estimate with
essentially no precision, and it is a different number for every seed.

\(s_*\) is not that. Equation (5) of the paper is

$$V(\hat\beta) = (X^{\mathsf T}\Sigma^{-1}X)^{-1},$$

and the paper prefaces it with the condition that carries all the weight:
"**provided that there is a postulated fixed value** for the variance–covariance
matrix". \(\Sigma\) is not estimated from the two people. It is supplied — from
stage one, by `slope_params()`, out of a previous dataset that may have had
hundreds of participants. Given \(\Sigma\), equation (5) is not an estimate of a
standard error; it is the exact sampling variance of a generalised-least-squares
estimator, an algebraic function of the design and the assumed variance
components. Nobody has to be able to fit anything.

So \(s_*\) is better read as a **design constant**: the standard error that
*would* attach to \(\hat\beta_2\) in a trial of one participant per arm, if the
variance components were known. It is a unit of information, not a measurement.
Its own uncertainty — the fact that stage one only estimated \(\Sigma\) — is a
separate matter, and is what the bootstrap in section 2.6 of the paper is for.

## 2. Why two people, and why one in each arm

"Two-person trial" reads like an arbitrary choice of a small number. It is not
arbitrary: two is the smallest trial in which \(\beta_2\) exists at all, and the
allocation is forced.

Information from independent participants adds. For participant \(i\) with
design matrix \(X_i\), the contribution to \(X^{\mathsf T}\Sigma^{-1}X\) is
\(X_i^{\mathsf T}\Sigma^{-1}X_i\), and the total is the sum over participants —
this is just what the block-diagonal structure of the full covariance matrix
does when you multiply it out. So ask what a *single* participant contributes.

```{r ranks}
Si <- solve(Sigma)
X_ctrl <- cbind(1, visits, 0)        # g = 0: the interaction column is dead
X_trt  <- cbind(1, visits, visits)   # g = 1: it duplicates the time column

info <- function(X) crossprod(X, Si %*% X)
c(control_only = qr(info(X_ctrl))$rank,
  treated_only = qr(info(X_trt))$rank,
  one_of_each  = qr(info(X_ctrl) + info(X_trt))$rank)
```

Each participant alone gives a rank-2 information matrix for three parameters,
and for a different reason in each arm. A control participant has \(g_i = 0\),
so the third column of their design matrix is identically zero: their data
contain no information about \(\beta_2\) whatsoever. A treated participant has
\(g_i = 1\), so their third column *equals* their second: they cannot separate
\(\beta_1\) from \(\beta_2\), only the sum. Neither can be inverted, so neither
alone has a treatment-effect variance.

One of each is rank 3. That is the point. **Two is the minimum, and it must be
one per arm** — not because two-person trials are interesting, but because the
matched pair is the smallest self-contained replicate the design decomposes
into. \(N = 2\) is where the calculation *starts*, not a small case of it.

Note that a single follow-up visit is enough for this:

```{r rank-one-followup}
v1 <- c(0, 1)
S1 <- slope_sigma(pars, v1); S1i <- solve(S1)
qr(crossprod(cbind(1, v1, 0), S1i %*% cbind(1, v1, 0)) +
   crossprod(cbind(1, v1, v1), S1i %*% cbind(1, v1, v1)))$rank
```

Four observations, three parameters, one residual degree of freedom — and it
does not matter, because \(\Sigma\) is not being estimated. What *would* fail is
a trial with a baseline visit only: then the time column is zero and there is no
slope to compare. This is exactly why the Dawson–Lagakos stratum of
baseline-only dropouts is discarded in `trial_design()` rather than given a
weight.

## 3. The \(\sqrt{N}\) scaling, derived

Now the second assertion. Take \(N\) participants per arm, all independent, all
on the same visit schedule. Stack them: \(X\) is the \(2N(m{+}1) \times 3\)
matrix of their design matrices piled up, and the full covariance is
block-diagonal with \(2N\) copies of \(\Sigma\) on the diagonal. A
block-diagonal matrix inverts blockwise, so

$$X^{\mathsf T}\Sigma_{\text{full}}^{-1}X \;=\; \sum_{i=1}^{2N} X_i^{\mathsf T}\Sigma^{-1}X_i \;=\; N\left(X_{\text{ctrl}}^{\mathsf T}\Sigma^{-1}X_{\text{ctrl}} + X_{\text{trt}}^{\mathsf T}\Sigma^{-1}X_{\text{trt}}\right) \;=\; N \, \mathcal{I}_2,$$

where \(\mathcal{I}_2\) is the pair's information matrix. Every participant in
an arm contributes the *same* matrix, because they share a schedule and a group
indicator, so the sum is a count. Then

$$V(\hat\beta) = (N\mathcal{I}_2)^{-1} = \tfrac{1}{N}\,\mathcal{I}_2^{-1}, \qquad\text{so}\qquad V(\hat\beta_2) = \frac{(\mathcal{I}_2^{-1})_{33}}{N} = \frac{s_*^2}{N}.$$

That is the whole derivation. It is an **exact identity**, not an asymptotic
approximation: scalar multiples pull straight through a matrix inverse, so no
limit is taken and no normality is used beyond what already defines the GLS
variance. The paper's "it follows from standard theory" is doing no work that
these three lines do not do.

There is one objection worth answering, because it is the reason the result
looks less trivial than it is. The pairs are *not* separate analyses:
\(\beta_0\) and \(\beta_1\) are common to everybody, so you cannot fit each pair
on its own and average \(N\) independent estimates of \(\beta_2\). But nothing
here does that. Information is summed across participants *before* the single
inversion at the end, and the nuisance parameters are handled once, jointly, in
that one inversion. It happens that the sum is \(N\mathcal I_2\), so the answer
coincides with what naive averaging would have given. That coincidence is a
consequence of every participant contributing an identical block; it is not a
premise.

Check it against the full \(2N(m{+}1)\)-row calculation:

```{r scaling-check}
N <- 7
X_big <- rbind(X_ctrl[rep(1:3, N), ], X_trt[rep(1:3, N), ])
S_big <- matrix(0, 6 * N, 6 * N)
for (i in seq_len(2 * N)) {
  k <- (i - 1) * 3 + 1:3
  S_big[k, k] <- Sigma
}
V_full <- solve(crossprod(X_big, solve(S_big) %*% X_big))

c(full_GLS = V_full[3, 3], s_star_sq_over_N = slope_var(pars, visits) / N)
```

Identical, to the last bit. The conditions this needs are worth naming, because
each is a real restriction on what `slopepower` can be used for:

* **Independence between participants.** Any further level of clustering — site,
  therapist, family — breaks the block-diagonal structure and the sum is no
  longer \(N\mathcal I_2\). The package has no facility for this and the
  `DESCRIPTION` says so.
* **A common schedule.** Every participant contributes the same \(X_i\) within
  arm. Dropout violates this, which is why it is handled by a separate device;
  see section 8.
* **1:1 allocation.** \(s_*/\sqrt N\) counts \(N\) *pairs*. Under \(r{:}1\)
  allocation the two arms contribute different counts and the pair is no longer
  the replicate; `slope_power()` enforces an even \(n\) split evenly for exactly
  this reason.
* **\(\Sigma\) the same in both arms and known.** The first is the paper's model
  assumption (variance components are constrained equal across arms in stage
  one); the second is section 1 above.

Finally, a notational trap, since it costs people a factor of two. In equation
(6) of the paper, \(N\) is the number **per arm** — it is the \(N\) of
\(s_*/\sqrt N\). The command's printed output labels the *total* as `N` and the
per-arm figure as `N per arm`:

```{r n-convention}
p1 <- slope_params(sdmt ~ visit | id, data = slpower1)
ss <- slope_sample_size(p1, c(0, 1, 2), effectiveness = 0.33)
z  <- qnorm(0.975) + qnorm(0.8)
c(eq6            = ceiling((z * sqrt(slope_var(p1, c(0, 1, 2))) / (0.33 * abs(p1$slope)))^2),
  reported_n_arm = ss$n / 2,
  reported_n     = ss$n)
```

## 4. A closed form the paper does not give

The information matrix of a pair has enough structure to invert by hand. Write
the three GLS inner products

$$A = u^{\mathsf T}\Sigma^{-1}u, \qquad B = u^{\mathsf T}\Sigma^{-1}t, \qquad C = t^{\mathsf T}\Sigma^{-1}t.$$

With \(X_{\text{ctrl}} = [\,u \; t \; 0\,]\) and \(X_{\text{trt}} = [\,u \; t \; t\,]\),

$$\mathcal I_2 = \begin{pmatrix} A & B & 0 \\ B & C & 0 \\ 0 & 0 & 0\end{pmatrix} + \begin{pmatrix} A & B & B \\ B & C & C \\ B & C & C\end{pmatrix} = \begin{pmatrix} 2A & 2B & B \\ 2B & 2C & C \\ B & C & C\end{pmatrix}.$$

The quantity we want is \((\mathcal I_2^{-1})_{33}\), which is the reciprocal of
the Schur complement of the leading \(2\times 2\) block:

$$s_*^{-2} \;=\; C - \begin{pmatrix} B & C\end{pmatrix}\begin{pmatrix} 2A & 2B \\ 2B & 2C\end{pmatrix}^{-1}\begin{pmatrix} B \\ C\end{pmatrix}.$$

Now
\(\bigl(\begin{smallmatrix} A & B \\ B & C\end{smallmatrix}\bigr)^{-1}\bigl(\begin{smallmatrix} B \\ C\end{smallmatrix}\bigr) = \frac{1}{AC-B^2}\bigl(\begin{smallmatrix} C & -B \\ -B & A\end{smallmatrix}\bigr)\bigl(\begin{smallmatrix} B \\ C\end{smallmatrix}\bigr) = \frac{1}{AC-B^2}\bigl(\begin{smallmatrix} 0 \\ AC-B^2\end{smallmatrix}\bigr) = \bigl(\begin{smallmatrix} 0 \\ 1\end{smallmatrix}\bigr)\),
so the correction term is \(\tfrac12 (B, C)(0,1)^{\mathsf T} = C/2\), and

$$\boxed{\;s_*^2 \;=\; \frac{2}{\,t^{\mathsf T}\Sigma^{-1}t\,}\;}$$

\(A\) and \(B\) cancel completely. The intercept, and everything the baseline
measurement's own variance contributes through \(A\), leave no trace in
\(s_*\).

```{r closed-form}
closed <- function(pars, v) 2 / drop(t(v) %*% solve(slope_sigma(pars, v)) %*% v)

designs <- list(c(0, 1), c(0, 1, 2), c(0, 1, 2, 3), seq(0, 2, by = 0.5), c(0, 2, 5, 7.3))
t(vapply(designs, function(v) c(slope_var = slope_var(pars, v), closed_form = closed(pars, v)),
         numeric(2)))
```

### What the formula says

\(t^{\mathsf T}\Sigma^{-1}t\) has a variational characterisation. For any
non-zero contrast vector \(c\) over the visits, Cauchy–Schwarz in the
\(\Sigma\) inner product gives

$$\max_{c \neq 0} \frac{(c^{\mathsf T}t)^2}{c^{\mathsf T}\Sigma c} \;=\; t^{\mathsf T}\Sigma^{-1}t, \qquad \text{attained at } c^* \propto \Sigma^{-1}t.$$

Read that backwards and \(s_*^2 = 2/C\) becomes a statement about an estimator.
Apply the *same* contrast \(c\) to each participant's measurements and take the
difference between the two arms. The control participant's expectation is
\((c^{\mathsf T}u)\beta_0 + (c^{\mathsf T}t)\beta_1\) and the treated
participant's is \((c^{\mathsf T}u)\beta_0 + (c^{\mathsf T}t)(\beta_1 + \beta_2)\),
so differencing cancels \(\beta_0\) and \(\beta_1\) exactly, whatever \(c\) is,
and leaves \((c^{\mathsf T}t)\beta_2\). Hence

$$\tilde\beta_2(c) = \frac{c^{\mathsf T}(y_{\text{trt}} - y_{\text{ctrl}})}{c^{\mathsf T}t}, \qquad \operatorname{Var}\tilde\beta_2(c) = \frac{2\,c^{\mathsf T}\Sigma c}{(c^{\mathsf T}t)^2},$$

and minimising over \(c\) gives exactly \(2/C\). So the GLS estimator from a
pair *is* a between-person difference of one optimally weighted within-person
contrast — not an approximation to one:

```{r contrast}
cstar <- solve(Sigma) %*% visits

set.seed(4)
y_ctrl <- rnorm(3); y_trt <- rnorm(3)          # arbitrary data
X_pair <- rbind(X_ctrl, X_trt)
S_pair <- rbind(cbind(Sigma, 0 * Sigma), cbind(0 * Sigma, Sigma))
beta_gls <- solve(crossprod(X_pair, solve(S_pair) %*% X_pair),
                  crossprod(X_pair, solve(S_pair) %*% c(y_ctrl, y_trt)))

c(gls = beta_gls[3],
  contrast = drop(t(cstar) %*% (y_trt - y_ctrl)) / drop(t(cstar) %*% visits))
```

Two consequences follow immediately, neither stated in the paper.

**Adding a visit can never hurt.** Enlarging the schedule enlarges the set of
available contrasts — pad any old \(c\) with a zero in the new coordinate and it
gives the same value — so the maximum cannot fall, \(C\) cannot fall, and
\(s_*^2\) cannot rise. That is Table 1 of the paper, proved rather than
observed. (It says nothing about *how much* a visit helps, which can be
almost nothing — see the `m1` and `m2` columns in section 6.)

```{r monotone}
set.seed(2)
worse <- 0
for (k in 1:2000) {
  base  <- sort(c(0, runif(sample(1:4, 1), 0.1, 6)))
  extra <- runif(1, 0.1, 6)
  if (slope_var(pars, sort(c(base, extra))) > slope_var(pars, base) + 1e-10) worse <- worse + 1
}
worse
```

**A common baseline mean is worth something specific.** Suppose you refused to
assume the two arms share \(\beta_0\) and gave each participant their own
intercept. Each person's own slope would then have variance \(A/(AC - B^2)\),
and differencing two of them gives \(2A/(AC - B^2)\). Against \(s_*^2 = 2/C\)
the ratio is \(1/(1 - B^2/AC)\): the randomisation assumption that lets the arms
share an intercept is worth exactly that factor.

```{r shared-intercept}
gain <- function(pars, v) {
  Si <- solve(slope_sigma(pars, v)); u <- rep(1, length(v))
  A <- drop(t(u) %*% Si %*% u); B <- drop(t(u) %*% Si %*% v); C <- drop(t(v) %*% Si %*% v)
  c(shared = 2 / C, separate = 2 * A / (A * C - B^2), ratio = 1 / (1 - B^2 / (A * C)))
}
t(vapply(list(c(0, 1), c(0, 1, 2), c(0, 1, 2, 3), c(0, 2, 5, 7.3)),
         function(v) gain(pars, v), numeric(3)))
```

It is small here, and it happens to vanish at `c(0, 1, 2)` where \(B = 0\) —
a coincidence of this particular parameter set and schedule, not a rule.

## 5. Baseline is a covariate, and the formula says so

Because \(t_0 = 0\), the index \(j = 0\) drops out of
\(t^{\mathsf T}\Sigma^{-1}t = \sum_{j,k} t_j t_k (\Sigma^{-1})_{jk}\)
entirely. What survives is the follow-up block of \(\Sigma^{-1}\) — and by the
block-inverse formula, that block is

$$\bigl[(\Sigma^{-1})_{ff}\bigr]^{-1} = \Sigma_{ff} - \Sigma_{f0}\Sigma_{00}^{-1}\Sigma_{0f} = \operatorname{Var}(y_f \mid y_0),$$

the covariance of the follow-up measurements **conditional on baseline**. So

$$s_*^2 \;=\; \frac{2}{\,t_f^{\mathsf T}\,\operatorname{Var}(y_f \mid y_0)^{-1}\,t_f\,}.$$

```{r baseline-cov}
via_conditional <- function(pars, v) {
  S <- slope_sigma(pars, v)
  Vc <- S[-1, -1, drop = FALSE] - S[-1, 1, drop = FALSE] %*%
        solve(S[1, 1, drop = FALSE]) %*% S[1, -1, drop = FALSE]
  2 / drop(t(v[-1]) %*% solve(Vc) %*% v[-1])
}
t(vapply(designs, function(v) c(slope_var = slope_var(pars, v),
                                conditional = via_conditional(pars, v)), numeric(2)))
```

This is the ANCOVA connection the paper's introduction gestures at — "the
baseline measure of the outcome \(y_{i0}\) is treated as a correlated outcome",
and the model "will give the same estimated treatment effect as analysis of
covariance" — made explicit and exact. Baseline enters \(s_*\) in precisely the
way a covariate does: not by supplying a data point about the slope (its time is
zero, so it carries none), but by shrinking the covariance of the points that
do. Its whole contribution is the amount \(\Sigma_{f0}\Sigma_{00}^{-1}\Sigma_{0f}\)
it removes, which is driven by \(\sigma_a^2\) and \(\sigma_{ab}\) and destroyed
by measurement error \(\sigma_\epsilon^2\).

## 6. One follow-up visit, in closed form — and a floor on sample size

With \(m = 1\) and visits \((0, t)\) everything is \(2\times 2\) and
\(s_*^2 = 2\det\Sigma / (t^2\Sigma_{00})\) unwinds to

$$s_*^2 \;=\; \underbrace{\frac{2\sigma_\epsilon^2(2\sigma_a^2 + \sigma_\epsilon^2)}{t^2(\sigma_a^2 + \sigma_\epsilon^2)}}_{\text{shrinks like } t^{-2}} \;+\; \underbrace{\frac{4\sigma_\epsilon^2\sigma_{ab}}{t(\sigma_a^2+\sigma_\epsilon^2)}}_{\text{shrinks like } t^{-1}} \;+\; \underbrace{2\left(\sigma_b^2 - \frac{\sigma_{ab}^2}{\sigma_a^2+\sigma_\epsilon^2}\right)}_{\text{constant in } t}.$$

```{r one-followup}
one_fu <- function(p, t) {
  P <- p$sigma2_intercept + p$sigma2_residual
  2 * p$sigma2_residual * (2 * p$sigma2_intercept + p$sigma2_residual) / (t^2 * P) +
    4 * p$sigma2_residual * p$sigma_cov / (t * P) +
    2 * (p$sigma2_slope - p$sigma_cov^2 / P)
}
tt <- c(0.5, 1, 2, 5, 20, 1000)
rbind(t = tt,
      slope_var = vapply(tt, function(t) slope_var(pars, c(0, t)), 0),
      closed    = one_fu(pars, tt))
```

The third term does not involve \(t\). **Lengthening a two-visit trial has a
hard floor**, and it is not zero: however far apart you place the two visits,
\(s_*^2\) cannot fall below \(2(\sigma_b^2 - \sigma_{ab}^2/(\sigma_a^2+\sigma_\epsilon^2))\).
The reason is not subtle once seen — with two visits you learn each
participant's own slope, but each participant's own slope *is* \(\beta_1 +
\beta_2 g_i + b_i\), and the \(b_i\) never go away. You are estimating a
difference of means of a quantity with irreducible between-person variance
\(\sigma_b^2\), less whatever the baseline measurement told you about \(b_i\)
through \(\sigma_{ab}\).

Adding interim visits relaxes the floor, but only to a point. Densifying the
schedule drives \(\sigma_\epsilon^2\) out of the problem entirely, and the limit
is the same expression with \(\sigma_\epsilon^2\) set to zero:

$$s_*^2 \;\longrightarrow\; 2\left(\sigma_b^2 - \frac{\sigma_{ab}^2}{\sigma_a^2}\right) \qquad\text{as the schedule becomes long and dense.}$$

```{r floor}
c(m1 = slope_var(pars, c(0, 3)),      m2 = slope_var(pars, seq(0, 3, length.out = 3)),
  m6 = slope_var(pars, seq(0, 3, length.out = 7)),
  dense_3yr  = slope_var(pars, seq(0, 3,  length.out = 401)),
  dense_20yr = slope_var(pars, seq(0, 20, length.out = 2001)),
  floor = 2 * (pars$sigma2_slope - pars$sigma_cov^2 / pars$sigma2_intercept))
```

Because \(N \propto s_*^2\), this is a floor on sample size that no design can
beat. For the fitted `slpower1` parameters and the paper's 33% effectiveness
target:

```{r floor-n}
b2 <- 0.33 * abs(p1$slope)
floor_var <- 2 * (p1$sigma2_slope - p1$sigma_cov^2 / p1$sigma2_intercept)

c(floor_n_total  = 2 * ceiling(z^2 * floor_var / b2^2),
  achieved_50yr  = slope_sample_size(p1, seq(0, 50, length.out = 501),
                                     effectiveness = 0.33)$n,
  paper_2yr      = ss$n)
```

A fifty-year trial with a visit every five weeks would still need about
`r 2 * ceiling(z^2 * floor_var / b2^2)` participants — and it reaches the floor
exactly, to the participant. That number is a property of the *disease*, through
\(\sigma_b^2\) and \(\sigma_{ab}/\sigma_a^2\), and no amount of trial design
touches it. It is a useful sanity check to compute before exploring schedules:
if the floor is already unaffordable, the design search is pointless and the
target effect size, not the visit schedule, is what has to change.

Because it is useful, the package computes it directly, and the two lines above
are what it does. `slope_var_floor()` is the limiting \(s_*^2\) and
`slope_sample_size_floor()` the sample size it implies:

```{r floor-fn}
c(hand = floor_var, slope_var_floor = slope_var_floor(p1))

flr <- slope_sample_size_floor(p1, effectiveness = 0.33)
c(floor = flr$n, paper_2yr = ss$n, ratio = ss$n / flr$n)
```

Given a result you already have, the floor for its own settings — same
effectiveness, same power, same alpha — comes from the result itself:

```{r floor-from-result}
slope_sample_size_floor(ss)$n
```

Neither takes a visit schedule, because neither depends on one. That is the
whole content of the bound.

Two warnings about reading it, both visible in the algebra above.

It is an infimum and not a minimum. Split the covariance as
\(\Sigma = \Sigma_0 + \sigma_\epsilon^2 I\), where \(\Sigma_0\) is the
noise-free part; then \(c^{\mathsf T}\Sigma c = c^{\mathsf T}\Sigma_0 c +
\sigma_\epsilon^2\lVert c\rVert^2\) is *strictly* larger for every \(c \neq 0\),
so the maximum in section 4's variational characterisation — and hence \(C\) —
is strictly smaller than its limit, and \(s_*^2 = 2/C\) strictly larger than the
floor. Every finite schedule is above it, however long and however dense; the
fifty-year design ties only after `ceiling()` rounds it to a whole participant.

And length alone will not get you there. The \(m = 1\) limit above is
\(2(\sigma_b^2 - \sigma_{ab}^2/(\sigma_a^2 + \sigma_\epsilon^2))\), not
\(2(\sigma_b^2 - \sigma_{ab}^2/\sigma_a^2)\):

```{r floor-two-visit}
c(two_visits_far_apart = slope_var(p1, c(0, 1e4)),
  two_visit_limit      = 2 * (p1$sigma2_slope -
                              p1$sigma_cov^2 / (p1$sigma2_intercept + p1$sigma2_residual)),
  floor                = slope_var_floor(p1))
```

A single distant follow-up tells you each participant's slope, but it corrects
for their intercept using one noisy baseline measurement. Only repeated
measurement recovers the whole of what \(\sigma_{ab}\) has to offer.

## 7. Does \(s_*/\sqrt N\) describe a real trial?

Everything so far is exact given \(\Sigma\). A real trial estimates \(\Sigma\)
too. The last check is empirical: simulate trials from the assumed model, fit
them the way they would actually be fitted — REML, variance components
estimated — and compare the spread of \(\hat\beta_2\) with \(s_*/\sqrt N\).

```{r simulation}
truth <- list(b0 = 34, b1 = -1.8, b2 = 0.6, sa = 100, sb = 2, sab = 5, se = 10)
N_arm <- 60
L <- chol(matrix(c(truth$sa, truth$sab, truth$sab, truth$sb), 2))

sim_trial <- function() {
  n <- 2 * N_arm
  re <- matrix(rnorm(2 * n), n, 2) %*% L
  id <- rep(seq_len(n), each = 3)
  tt <- rep(visits, n)
  gg <- rep(rep(0:1, each = N_arm), each = 3)
  y  <- truth$b0 + truth$b1 * tt + truth$b2 * gg * tt +
        re[id, 1] + re[id, 2] * tt + rnorm(n * 3, 0, sqrt(truth$se))
  data.frame(id = factor(id), t = tt, g = gg, y = y)
}

set.seed(20260810)
R <- 200
res <- vapply(seq_len(R), function(r) {
  f <- lme(y ~ t + g:t, random = ~ t | id, data = sim_trial(), method = "REML",
           control = lmeControl(opt = "optim"))
  summary(f)$tTable["t:g", c("Value", "Std.Error")]
}, numeric(2))

c(predicted_se  = sqrt(slope_var(pars, visits) / N_arm),
  empirical_sd  = sd(res["Value", ]),
  mean_model_se = mean(res["Std.Error", ]),
  mean_estimate = mean(res["Value", ]), truth = truth$b2)
```

The average standard error the mixed model reports lands within a fraction of a
percent of \(s_*/\sqrt{N}\), and the empirical standard deviation of the `r R`
fitted treatment effects is a few percent below it — comfortably inside the
Monte Carlo error of a standard deviation from `r R` replicates, which is about
\(1/\sqrt{2(R-1)} \approx\) `r round(100 / sqrt(2 * (R - 1)), 1)`%. So \(s_*\)
is not merely an internally consistent construction: it predicts what the
analysis of the planned trial will actually produce, and it does so without
being told the variance components the real analysis had to estimate for
itself.

## 8. Where the two-person device stops working

The construction assumed every participant contributes an identical
information block. Two things in the package break that assumption, and it is
worth being clear that they are handled *outside* \(s_*\), not inside it.

**Dropout.** Participants who withdraw have different schedules, so
\(X^{\mathsf T}\Sigma^{-1}X \neq N\mathcal I_2\) and there is no single \(s_*\).
The Dawson–Lagakos pattern-mixture device computes a separate \(s_*\) for each
dropout stratum — each with its own leading submatrix of \(\Sigma\), the
covariance being unaffected by which *other* visits exist — and combines them
on the \(1/N\) scale rather than the \(s_*\) scale, since sample size is what
is additive across strata:

```{r dropout}
design <- trial_design(c(0, 1, 2), dropout = c(0, 0.25))
strata <- c(completers = slope_var(pars, c(0, 1, 2)), withdrawers = slope_var(pars, c(0, 1)))
strata

# effective s*^2: reciprocal of the weighted mean of the reciprocals
c(effective = 1 / (0.75 / strata[["completers"]] + 0.25 / strata[["withdrawers"]]),
  reported  = slope_power(pars, design, effectiveness = 0.33, n = 400)$var_tte)
```

Note that the effective value is not between the strata in the way a naive
average would be — the weighting is harmonic, so the informative stratum
dominates. `slope_effect_size()` documents the same arithmetic on the effect
scale.

**Unequal allocation.** \(s_*/\sqrt N\) counts pairs, so there is no \(N\) to
divide by under \(r{:}1\) randomisation. The package does not offer it;
`slope_power()` rounds an odd `n` down to an even one and splits it evenly.

Everything else in the package — the effect size, the target treatment effect,
the two closed-form solutions for \(N\) and power — takes \(s_*\) as given. It
really is the single number the whole calculation turns on, which is why it
seemed worth thirty lines of algebra rather than two sentences.

## References

Dawson, J. D., and S. W. Lagakos. 1993. Size and power of two-sample tests of
repeated measures data. *Biometrics* 49: 1022–1032.

Frost, C., M. G. Kenward, and N. C. Fox. 2008. Optimizing the design of clinical
trials where the outcome is a rate. Can estimating a baseline rate in a run-in
period increase efficiency? *Statistics in Medicine* 27: 3717–3731.

Nash, S., K. E. Morgan, C. Frost, and A. Mulick. 2021. Power and sample-size
calculations for trials that compare slopes over time: Introducing the
slopepower command. *Stata Journal* 21(3): 575–601.
